### 2011-11-17 paper/finalproject.tex
- Committed final project LaTeX file using appropriate SIGPLAN   template referenced on course web page.

%-----------------------------------------------------------------------------
%               CSCI5525 Final Project LaTeX file.
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{graphicx}
\usepackage{listings}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage[pdfborder={0 0 0}]{hyperref}

\lstset{
basicstyle=\small,
numberstyle=\small,
commentstyle=\textit
stepnumber=1
}

\begin{document}

\conferenceinfo{CSCI5525 '11}{November 16th, 2011, Boulder,CO.} 
\copyrightyear{2011} 
\copyrightdata{[Like a Boss.]}

\titlebanner{Reference Counting in P3}        % These are ignored unless
\preprintfooter{An implementation of reference counting for P3}   % 'preprint' option specified.

\title{Reference Counting in P3}
%\subtitle{Subtitle Text, if any}

\authorinfo{Rob Elsner}
           {University of Colorado, Boulder}
           {rob.elsner@colorado.edu}
\authorinfo{Brent Smith}
           {University of Colorado, Boulder}
           {brent.m.smith@colorado.edu}

\maketitle

\begin{abstract}
Automatic memory management is a common feature in modern languages, eliminating the need to
manually allocate and deallocate memory at the proper program points.  This feature helps to
eliminate costly errors that may go undetected in short lived programs, but cause long running 
programs to eventually fail. 
In this paper, we will discuss the implementation of a reference counting
garbage collector for a subset of the Python programming language.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
memory management, garbage collector

\keywords
keyword1, keyword2

\section{Introduction}

There are currently two widely used strategies to perform automatic memory management.  We will
discuss both strategies briefly, consider the tradeoffs associated with each, and then discuss
how the reference counting scheme fits in to the current architecture of our compiler.

Tracing garbage collection is ....
%ref{mccarthy}
Tracing garbage collection is commonly associated with the garbage collector implementation  
provided by the Java programming language, but many languages have libraries that provide GC 
implementations 
%\ref{c++ implementations}
or allow the user the option of manual memory management or a tracing GC.
%\ref{Apple's objective C framework}

Reference counting \cite{Collins:1960} is way cooler since you don't suffer long pause times, but 
may not achieve equivalent throughput to tracing garbage collection.  It's also wicked hard 
to get right on multi-threaded contexts.  You have to be an evil genious to get this shit right,
and use atomic compare and swap operations, whatever those are.

\begin{figure}
\normalsize % use normal font size in the figure
\input{slot_runtime.tex}
\caption{Benchmark results on Local Cluster}
\label{fig:slot_runtime}
\end{figure}

\appendix
\section{Appendix of Appendices}

Here we list the categories of topics discussing different types of thingies 
and such as, whatnot and therefore. Thank you.

\acks

I'd like to thank all the beautiful children.

% We recommend abbrvnat bibliography style.
\bibliographystyle{abbrvnat}
% Use bibtex instead of the other crap
\bibliography{finalproject}

% The bibliography should be embedded for final submission.
%\begin{thebibliography}{}
%\softraggedright
%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...
%\end{thebibliography}

\end{document}

### 2011-12-09 compiler/comp_util.py
- Delete *.s and *.o files as part of clean target in Makefile - Add 3 test cases:   * test/basic.py   * test/freevar.py   * test/nested.py - Modify comp_util to use repr(n) when printing an unhandled expressoin - Don't raise an exception in gcflattener.py if we see an IfExp; raise   instead in gcrefcount.py - Fix bug with visit_Return in gcflattener.py that was causing the expression   to be visited twice - Fix visit_CallFuncIndirect in gcflattener.py; it was incorrectly adding   variables to statement lists and other odd things.  Make it easier to   understand by using a for loop instead of

@@ -299,7 +299,7 @@ def getLocalAssigns(n):
     elif isinstance(n, Class):
         return set([n.name])
     else:
-        raise Exception('Unhandled expression: "%s"' % n)
+        raise Exception('Unhandled expression: "%s"' % repr(n))
 
 class CallFuncIndirect(Node):
     def __init__(self, node, args, star_args = None, dstar_args = None, lineno=None):

### 2011-12-09 compiler/gcflattener.py

@@ -33,6 +33,7 @@ class GCFlattener:
         meth = None
         meth_name = 'visit_'+node.__class__.__name__
         meth = getattr(self, meth_name, None)
+        self.log.info(meth_name)
         if not meth:
             raise Exception('Unknown node: %s method: %s' % (node.__class__, meth_name))
         return meth(node, *args, **kwargs)
@@ -134,7 +135,6 @@ class GCFlattener:
         return (Name(varname), [varname]+tmpvarleft+tmpvarright, lhsstmtlist + rhsstmtlist + [Assign([AssName(varname, 'OP_ASSIGN')], result)])
     
     def visit_IfExp(self, node, *args, **kwargs):
-        raise Exception("We shouldn't have an IfExp at this point.  It should have been changed to an If by our flattener.")
         # Go ahead and flatten all expressions, including the test expression, as well as the 
         # "then" and "else" expressions.
         vartes, testtmpvars, test = self.visit(node.test)
@@ -205,19 +205,24 @@ class GCFlattener:
         return (Name(varname), [varname]+exprtmplist+subtmplist, exprstmtlist + substmtlist + [Assign([AssName(varname,'OP_ASSIGN')], Subscript(exprvar, node.flags, subvar))])
     
     def visit_Return(self, node, *args, **kwargs):
-        x = self.visit(node.value)
         retvar, tmplist, retstmtlist = self.visit(node.value)
         return retstmtlist + cleanup_tempvars(tmplist) + [Return(retvar)]
     
     def visit_CallFuncIndirect(self, node, *args, **kwargs):
         self.log.debug('CallFuncIndirect: args: %s', node.args)
         nodevar, nodetmpvars, nodestmtlist = self.visit(node.node)
-        tuplelist = [self.visit(x) for x in node.args]
-        varargs = [x[0] for x in tuplelist]
-        tmpvars = reduce(lambda x,y: x+y, [y for x,z,y in tuplelist if y != []], []) 
-        varstmts = [x[1] for x in tuplelist]
+        varargs=[]
+        tmpvars=[]
+        varstmts=[]
+        for x in node.args:
+            a, b, c = self.visit(x)
+            varargs.append(a)
+            tmpvars.append(b)
+            varstmts.append(c)
+        tmpvars = reduce(lambda x,y: x+y, tmpvars, [])
+        varstmts = reduce(lambda x,y: x+y, varstmts, [])
         varname = self.varalloc.get_next_var()
[…]

### 2011-12-09 compiler/gcrefcount.py

@@ -7,6 +7,59 @@ from p3uniquifyvars import P3UniquifyVars
 from gcflattener import GCFlattener
 import operator
 
+def getLocals(n):
+    """
+    Returns the set of variables that are assigned to within the current scope,
+    ignoring assignments in nested scopes (functions).
+    """
+    if isinstance(n, Module):
+        return getLocals(n.node)
+    elif isinstance(n, Stmt):
+        assigns = [getLocals(x) for x in n.nodes]
+        return reduce(lambda x,y: x|y, assigns, set([]))
+    elif isinstance(n, Printnl):
+        return set([])
+    elif isinstance(n, Assign):
+        if isinstance(n.nodes[0], (Subscript,AssAttr)):
+            # assigning to a subscript of a variable does not constitute
+            # assigning to the variable itself.  Return empty set.
+            # ditto for an object attribute
+            return set([])
+        elif isinstance(n.nodes[0], AssName):
+            return set([n.nodes[0].name])
+        else:
+            raise Exception('Unhandled Assign case: %s' % n.nodes[0])
+    elif isinstance(n, Discard):
+        # there shouldn't ever be a case where there will be an assignment in a Discard
+        # since a Discard is, by definition, a statement that produced no assignments
+        return set([])
+    elif isinstance(n, Return):
+        return set([])
+    elif isinstance(n, Function):
+        # a function definition is equivalent to an assignment to a 
+        # variable with the same name as the function
+        # No need to recurse into the Function since the intent of the function
+        # is to only find assigments for the local scope
+        return set([n.name])
+    elif isinstance(n, If):
+        testset = getLocals(n.tests[0][0])
+        thenset = getLocals(n.tests[0][1])
+        elseset = getLocals(n.else_)
+        return testset | thenset | elseset
+    elif isinstance(n, While):
+        testset = getLocals(n.test[1])
+        bodyset = getLocals(n.body)
+        elseset = getLocals(n.else_) if n.else_ is not None and n.else_ != [] else set()
+        return testset | bodyset | elseset
+    elif isinstance(n, (Add,UnarySub,CallFunc,Const,Name,Or,And,IfExp,List,Dict,Compare,Not,Subscript,Lambda,CallFuncIndirect)):
+        # these are all expressions, so no assignments
+        return set([])
+    elif isinstance(n, Class):
+        return set([n.name])
+    else:
[…]

### 2011-12-09 compiler/test/basic.py

@@ -1,0 +1,4 @@
+a = []
+b = []
+a = b
+print a

### 2011-12-09 compiler/test/freevar.py

@@ -1,0 +1,4 @@
+x=2
+def f():
+    print x
+f()

### 2011-12-09 compiler/test/nested.py

@@ -1,0 +1,3 @@
+def f():
+    return lambda: 1 + 2
+f()

### 2011-12-09 compiler/p3closureconvert.py
- p3closureconvert.py   * Add wrapper, gcflattener and gcrefcount phases - p3explicate.py   * Fix explication of While.  Since we added a flatten stage before     explication, the structure of our While node is different.  Take this into     account during explication. - p3flattener.py   * Updated to conform to the changed structure of the While node.   * Added wrapper phase - p3freevars.py   * Updated to conform to the changed structure of the While node. - p3heapify.py   * Updated to conform to the changed structure of the While node.
Test status:
class_tests/   - 34 of 34 passed
our_hw5_tes

@@ -27,6 +27,9 @@ if __name__ == "__main__":
     import sys, compiler
     import logging.config
     from p3declassify import P3Declassify
+    from gcflattener import GCFlattener
+    from p3wrapper import P3Wrapper
+    from gcrefcount import GCRefCount
     from p3explicate import P3Explicate
     from p3uniquifyvars import P3UniquifyVars
     from p3heapify import P3Heapify
@@ -38,14 +41,20 @@ if __name__ == "__main__":
     for testcase in testcases:
         varalloc = VariableAllocator()
         declassify = P3Declassify(varalloc)
+        wrapper = P3Wrapper()
         unique = P3UniquifyVars()
+        gcflatten = GCFlattener(varalloc)
+        gcrefcount = GCRefCount(varalloc)
         explicator = P3Explicate(varalloc)
         heap = P3Heapify(explicator)
         closure = P3ClosureConversion(explicator, varalloc)
 
         ast = compiler.parseFile(testcase)
         ast = declassify.transform(ast)
+        ast = wrapper.transform(ast)
         ast = unique.transform(ast)        
+        ast = gcflatten.transform(ast)        
+        ast = gcrefcount.transform(ast)        
         ast = explicator.explicate(ast)
         ast = heap.transform(ast)
         astlist = closure.transform(ast)

### 2011-12-09 compiler/p3explicate.py

@@ -9,8 +9,19 @@ class P3Explicate(P2Explicate):
     def __init__(self, varalloc, handleLambdas=True):
         P2Explicate.__init__(self, varalloc, handleLambdas)
             
+    # XXX: Had to change this after we added a flatten phase before explication.
+    # Because of this change, we have a tuple for the "test" attribute, instead of
+    # an expression.   The first element of the tuple is the variable associated 
+    # with the test, while the second element in the tuple is the Stmt node. 
+    # Below, we maintain the tuple for the "test" attribute, but since we have to
+    # inject some IfExp logic, we change the first element of the tuple to a Let
+    # node.   The second element remains the same, but is the explicated version
+    # of the Stmt node.
     def visit_While(self, node, *args, **kwargs):
-        test  = self.visit(node.test)
+        test0  = self.visit(node.test[0])
+        test1  = self.visit(node.test[1])
+        self.log.info('visit_While: test0=%s' % test0)
+        self.log.info('visit_While: test1=%s' % test1)
         body  = self.visit(node.body)
         var = Name(self.varalloc.get_next_var())
         # explicate the test so we can compare it to 0
@@ -19,7 +30,7 @@ class P3Explicate(P2Explicate):
                   ProjectTo('bool',var),
                   ProjectTo('bool',CallFunc(Name('is_true'),[var]))
                 )
-        return  While(Let(var, test, ifexp), body, [], node.lineno)
+        return  While((Let(var, test0, ifexp),test1), body, [], node.lineno)
 
     def visit_If(self, node, *args, **kwargs):
         if len(node.tests) > 1:

### 2011-12-09 compiler/p3flattener.py

@@ -3,6 +3,7 @@ from compiler.ast import *
 from comp_util import *
 from p2flattener import P2Flattener
 from p3declassify import P3Declassify
+from p3wrapper import P3Wrapper
 from p3explicate import P3Explicate
 from p3uniquifyvars import P3UniquifyVars
 from p3heapify import P3Heapify
@@ -25,8 +26,14 @@ class P3Flattener(P2Flattener):
             #statement
             flatbody = self.flatten(node.body)
             #expression
-            var, flattest = self.flatten(node.test) 
-            return [While((var,Stmt(flattest)), flatbody, [], node.lineno)]   
+            # the first element in the tuple is an expression, so it needs flattened.
+            var0, flattest0 = self.flatten(node.test[0]) 
+            # the second selement is a stmt node.
+            stmt = self.flatten(node.test[1])
+            # the statements associated with the newly flattened expression (node.test[0]),
+            # have to be run AFTER the statements in node.test[1]
+            stmt.nodes = stmt.nodes + flattest0
+            return [While((var0,stmt), flatbody, [], node.lineno)]   
         elif isinstance(node, If):
             # flatten the "test" expression
             vartes, test = self.flatten(node.tests[0][0])
@@ -61,6 +68,7 @@ if __name__ == "__main__":
         varalloc = VariableAllocator()
         declassify = P3Declassify(varalloc)
         unique = P3UniquifyVars()
+        wrapper = P3Wrapper()
         gcflatten = GCFlattener(varalloc, True)
         gcrefcount = GCRefCount(varalloc)
         explicator = P3Explicate(varalloc, handleLambdas=False)
@@ -70,6 +78,7 @@ if __name__ == "__main__":
 
         ast = compiler.parseFile(testcase)
         ast = declassify.transform(ast)
+        ast = wrapper.transform(ast)
         ast = unique.transform(ast)
         ast = gcflatten.transform(ast)        
         ast = gcrefcount.transform(ast)        

### 2011-12-09 compiler/p3freevars.py

@@ -27,7 +27,7 @@ class P3FreeVars(P2FreeVars):
     
     def visit_While(self, node, *args, **kwargs):
         self.log.debug('visit_While %s'%node)
-        test_b, test_f = self.visit(node.test)
+        test_b, test_f = self.visit(node.test[1])
         then_b, then_f = self.visit(node.body)
         return (set([]), test_f | then_f )
 

### 2011-12-09 compiler/p3heapify.py

@@ -20,7 +20,7 @@ class P3Heapify(P2Heapify):
 
     def getLambdaFreeVars(self, n):
         if isinstance(n, (While)):
-            test_fv = self.getLambdaFreeVars(n.test)
+            test_fv = self.getLambdaFreeVars(n.test[1])
             body_fv = self.getLambdaFreeVars(n.body)
             return test_fv | body_fv
         elif isinstance(n, If):
@@ -34,7 +34,7 @@ class P3Heapify(P2Heapify):
             #return super(P3Heapify, self).getLambdaFreeVars(n)
     
     def visit_While(self, node):
-        return While(self.visit(node.test), self.visit(node.body), [], node.lineno)
+        return While((self.visit(node.test[0]),self.visit(node.test[1])), self.visit(node.body), [], node.lineno)
 
     def visit_If(self, node):
         return If([(self.visit(node.tests[0][0]), self.visit(node.tests[0][1]))], self.visit(node.else_))

### 2011-12-10 compiler/gcrefcount.py
Fix list_add in runtime.c to call inc_ref on each element because it is creating a new list. 	dec_ref would be called on lhs and rhs of the Add, but since we were creating a brand new 	list we needed to track these new references since dec_ref would be called on the resulting 	object from the assignment

@@ -21,9 +21,9 @@ def getLocals(n):
         return set([])
     elif isinstance(n, Assign):
         if isinstance(n.nodes[0], (Subscript,AssAttr)):
-            # assigning to a subscript of a variable does not constitute
-            # assigning to the variable itself.  Return empty set.
-            # ditto for an object attribute
+             ##assigning to a subscript of a variable does not constitute
+             ##assigning to the variable itself.  Return empty set.
+             ##ditto for an object attribute
             return set([])
         elif isinstance(n.nodes[0], AssName):
             return set([n.nodes[0].name])

### 2011-12-10 compiler/runtime.c

@@ -16,6 +16,7 @@ static void print_float(double in);
 static void print_list(pyobj pyobj_list);
 static void print_dict(pyobj dict);
 static list list_add(list x, list y);
+static void print_pyobj(pyobj x);
 
 int tag(pyobj val) {
     return val & MASK;
@@ -140,6 +141,36 @@ static void print_bool(int b) {
         printf ("False");
 }
 
+static void print_class(pyobj obj) {
+    big_pyobj* b = project_big(obj);
+    printf("class with parents ");
+    print_pyobj(b->u.cl.parents);
+}
+
+static void print_object(pyobj obj) {
+    big_pyobj* b = project_big(obj);
+    printf("object of class ");
+    print_pyobj(b->u.obj.clazz);
+}
+
+static void print_ubmethod(pyobj obj) {
+    big_pyobj* b = project_big(obj);
+    printf("unbound method class ");
+    print_pyobj(b->u.ubm.clazz);
+}
+
+static void print_bmethod(pyobj obj) {
+    big_pyobj* b = project_big(obj);
+    printf("bound method receiver ");
+    print_pyobj(b->u.bm.receiver);
+    printf("    and fun at address %X\n", b->u.bm.fun);
+}
+
+static void print_fun(pyobj obj) {
+    big_pyobj* b = project_big(obj);
+    printf("function at address %X\n", b->u.f.function_ptr);
+}
+
 static void print_pyobj(pyobj x) {
     switch (tag(x)) {
     case INT_TAG:
@@ -160,7 +191,23 @@ static void print_pyobj(pyobj x) {
         case LIST:
             print_list(x);
             break;
+        case FUN:
+            print_fun(x);
+            break;
+        case CLASS:
+            print_class(x);
+            break;
+        case OBJECT:
+            print_object(x);
+            break;
+        case UBMETHOD:
+            print_ubmethod(x);
+            break;
+        case BMETHOD:
+            print_bmethod(x);
+            break;
         default:
+            printf("unknown tag for print_pyobj %d\n", tag(x));
             assert(0);
         }
         break;
@@ -619,10 +666,14 @@ static list list_add(list a, list b)
     c.len = a.len + b.len;
     c.data = (pyobj*)pymem_new(LIST, sizeof(pyobj) * c.len);
     int i;
-    for (i = 0; i != a.len; ++i)
[…]

### 2011-12-10 compiler/test/easy.py

@@ -1,0 +1,5 @@
+sieve = []
+
+sieve = sieve + [[1, True]]
+
+print sieve

### 2011-12-10 compiler/test/ref3.py

@@ -1,0 +1,4 @@
+test = [] + []
+input()
+test = 0
+input()

### 2011-12-10 compiler/test/ref4.py

@@ -1,0 +1,5 @@
+if input():
+	a = 2
+else:
+	b = 4
+print a

### 2011-12-10 compiler/test/ref5.py

@@ -1,0 +1,3 @@
+def f():
+    x = []
+f()

### 2011-12-10 compiler/test/sfun.py

@@ -1,0 +1,7 @@
+def m(x):
+    return 1
+
+def fact(x):
+    return m(x)
+
+print fact(5)

### 2011-12-10 compiler/test/sieve.py

@@ -26,6 +26,8 @@ i = 0
 while less(i,n):
 	sieve = sieve + [[i, True]]
 	i = i + 1
+	print i
+	print sieve
 
 nover2 = div(n, 2)
 i = 2

### 2011-12-12 paper/paper.tex
Adding latex paper

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage[pdftex]{graphicx}

\begin{document}

\conferenceinfo{WXYZ '05}{date, City.} 
\copyrightyear{2011} 
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Python to Assembly Compilation}
\subtitle{With Reference Counting Garbage Collection}

\authorinfo{Brent Smith}
           {University of Colorado, Boulder}
           {brent.m.smith@colorado.edu}
\authorinfo{Robert Elsner}
           {University of Colorado, Boulder}
           {robert.elsner@colorado.edu}

\maketitle

\begin{abstract}
Integrating a modern dynamic language in to existing systems-level languages (Assembly or C) easily has a number of challenges.  This paper will focus on automatic garbage collection within a subset of the Python language, to enable a Python-like program the ability to cleanly interact with Assembly or C routines without extra intervention from the software developer.  This paper will focus on reference counting garbage collection at the compiler level, to enable code injection which automatically frees non-referenced objects at the end of their useful lifetime.  We will also outline the analysis steps required in the compiler which enable reference tracking, how the abstract syntax tree of the program is altered and what trade-offs are made with such altercations.
\end{abstract}

\category{D.1.5}{Programming Techniques}{Object-oriented Programming}
\category{D.3.3}{Programming Languages}{Language Constructs And Features-Dynamic Storage Management}
\category{D.3.4}{Programming Languages}{Processors-Memory Management (garbage collection)}
\category{D.4.2}{Operating Systems}{Storage Management-Garbage collection}

\terms
Management, Performance, Design, Languages, Algorithms

\keywords
memory management, garbage collection, reference counting, compilation, compilers, programming languages

\section{Introduction}

Python is a modern high level language which has a reference counting garbage collector at its core.  It is also, by default, an interpreted language.  In many cases the usefulness of Python as a productive language is highly desired and a trade-off is made in terms of performance.  While numerous optimizations have been made to the Python run time there is still a usefulness to compile a high level language directly to machine code.  Most high level languages do not allow for direct memory management, instead leaving allocation and de-allocation of memory to the interpreter and garbage collection process.
There are two main classifications for garbage collectors: generational or tracing collectors and reference-counting collectors.  Generational or tracing collectors make periodic inspection of all objects to determine if they are reachable by a pointer or object reference.  These are the modern batch of garbage collectors and have the best performance 
Given a naive implementation of a direct Python to x86-assembly compiler which disregards all memory management, we explore the choices and consequences of adding a reference counting garbage collector.  Our base Python-like language includes classes, objects, functions, lambdas, simple math, lists, dictionaries and integer primitives.
\subsection{Disadvantages}
In general a reference count garbage collector will have shorter pause times for the garbage collection phase, but will incur a higher performance penalty \cite{joisha}\cite{blackburn}.  Reference-counting collectors have been implemented which defer the de-allocation, called lazy reference-counting, to some later time and some improvements on such strategies have been made \cite{boehm}.  Without extra work a strictly reference-counting collector will leak memory when references contain cycles, and multiple algorithms have been presented as solutions to this problem.
\subsection{Advantages}
One potential advantage that reference-counting garbage collectors have is the ability to be implemented in hardware \cite{joao}.  Given the significant cost of garbage collection \cite{hertz} any hardware assisted acceleration would be highly desirable if it is flexible enough to adapt across multiple garbage collection techniques. 
Real-time systems are rarely implemented in languages which are garbage collected, but reference-counting garbage collection has been adapted to hard real-time systems
\section{Implementation}
We develop a base compiler which generates x86 assembly code and then alter the compiler to include reference counting and garbage collection.
\subsection{Language}
A subset of Python has been chosen which includes functions, objects, lambdas, and the primitives integer, boolean.  We implement this as a reference compiler which is extended to include reference counting.  This language has no built in parallelism but Levanoni, et. al. have shown that reference counting is a valid approach for multprocessor systems.\cite{levanoni}
\par
Starting from our reference compiler, we treat any object which is not an integer or boolean primitive as a C-struct.  This structure was modified to include a reference counter and in all places where such a structure was allocated we converted the allocation to use a statistics tracking allocator and set the reference count to an initial value of 0.
We chose to implement our own memory routines primarily to investigate object lifetime (initial allocation time and final de-allocation time) as well as allow us the ability to detect memory leaks.  Since every allocation of an object in our language uses this wrapper we know exactly how much memory has been allocated and what objects were never de-allocated.
\begin{figure}
\begin{center}
\includegraphics[scale=0.48]{compiler_flow.png}
\end{center}
\caption{Compiler Architecture}
\label{fig-comparch}
\end{figure}

\subsection{Compiler Flow}
Refer to Figure~\ref{fig-comparch} for the compiler data flow.  Each stage operates on an abstract syntax tree, the relevant new stages are the garbage collection flattening stage and the reference count code injection stage.

\subsection{Syntax Tree Modifications}
We added two new stages to our reference compiler.  The first stage is a modified flattening stage, which 
\subsection{Runtime Modifications}
\subsection{Assembly Output}

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

We would like to thank Jeremy Siek and Evan Chang for the course notes from which the compiler design is derived.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright
\bibitem{joisha}
Pramod G. Joisha. 2006. Compiler optimizations for nondeferred reference: counting garbage collection. In Proceedings of the 5th international symposium on Memory management (ISMM '06). ACM, New York, NY, USA, 150-161. DOI=10.1145/1133956.1133976

\bibitem{blackburn}
Stephen M. Blackburn and Kathryn S. McKinley. 2003. Ulterior reference counting: fast garbage collection without a long wait. In Proceedings of the 18th annual ACM SIGPLAN conference on Object-oriented programing, systems, languages, and applications (OOPSLA '03). ACM, New York, NY, USA, 344-358. DOI=10.1145/949305.949336

\bibitem{boehm}
Hans-J. Boehm. 2004. The space cost of lazy reference counting. In Proceedings of the 31st ACM SIGPLAN-SIGACT symposium on Principles of programming languages (POPL '04). ACM, New York, NY, USA, 210-219. DOI=10.1145/964001.964019

\bibitem{joao}
José A. Joao, Onur Mutlu, and Yale N. Patt. 2009. Flexible reference-counting-based hardware acceleration for garbage collection. In Proceedings of the 36th annual international symposium on Computer architecture (ISCA '09). ACM, New York, NY, USA, 418-428. DOI=10.1145/1555754.1555806

\bibitem{hertz}
Matthew Hertz and Emery D. Berger. 2005. Quantifying the performance of garbage collection vs. explicit memory management. In Proceedings of the 20th annual ACM SIGPLAN conference on Object-oriented programming, systems, languages, and applications (OOPSLA '05). ACM, New York, NY, USA, 313-326. DOI=10.1145/1094811.1094836

\bibitem{levanoni}
Yossi Levanoni and Erez Petrank. 2006. An on-the-fly reference-counting garbage collector for java. ACM Trans. Program. Lang. Syst. 28, 1 (January 2006), 1-69. DOI=10.1145/1111596.1111597

\end{thebibliography}

\end{document}

### 2011-12-12 paper/paper.tex
added some more to the implementation description

@@
 Given the significant cost of garbage collection \cite{hertz} any hardware assisted acceleration would be highly desirable if it is flexible enough to adapt across multiple garbage collection techniques.
-Real-time systems are rarely implemented in languages which are garbage collected, but reference-counting garbage collection has been adapted to hard real-time systems
+\par
+Real-time systems are rarely implemented in languages which are garbage collected, but reference-counting garbage collection has been adapted to hard real-time systems \cite{?}.
+Due to immediate knowledge that an object is no longer in use reference counting garbage collection can allow for destructors or finalizers to be run immediately which can enhance the clean up of system resources.
 \section{Implementation}
@@
 \par
-Starting from our reference compiler, we treat any object which is not an integer or boolean primitive as a C-struct.
+Starting from our reference compiler, we treat any object which is not an integer or boolean primitive as a C structure.
 This structure was modified to include a reference counter and in all places where such a structure was allocated we converted the allocation to use a statistics tracking allocator and set the reference count to an initial value of 0.
+\par
 We chose to implement our own memory routines primarily to investigate object lifetime (initial allocation time and final de-allocation time) as well as allow us the ability to detect memory leaks.
 Since every allocation of an object in our language uses this wrapper we know exactly how much memory has been allocated and what objects were never de-allocated.
+\par
+The basic premise of such routines is to wrap the provided malloc and free, request slightly more memory than required.
+That extra memory is what where the reference counter is stored and the pointer location is returned which skips the counter.
+The memory routines also create a linked list to track all of the allocations in a structure for statistics gathering and leak detection.
 \begin{figure}
@@
 \subsection{Syntax Tree Modifications}
-We added two new stages to our reference compiler.  The first stage is a modified flattening stage, which 
+We added two new stages to our reference compiler.
+The first stage is a modified flattening stage, which produces a flattened AST to aid in the reference determination.
+The next stage is the actual reference determination and code injection, where the compiler determines what variables will be objects and when assignments happen and adds appropriate calls to increment and decrement the reference counter of the object as needed.
+\par
+Our initial assumption was that we could easily add the increment and decrement reference counter operations in the reference compilers flatten stage, however because this stage is after type explication and there are variables created which do not correspond to a type in the language the pre-flatten stage was necessary.
+This stage converts all expressions in to assignments to temporary variables.
+At this stage we know every expression is operating on types.
+\par
+The next new stage is the reference counting code injection.
+In this stage we take the flattened tree and look for assignments which are setting list elements, dictionary values, class attributes, object attributes or variables.
+The compiler determines what variables are created in the current scope and injects code at the beginning of that scope to set the value initially.
+We decided on this approach to simplify determining when an object is first assigned and this allows us to simply decrement the reference counter to the object before an assign, regardless of the variables state.
 \subsection{Runtime Modifications}
+Our runtime consisted of C functions which took care of allocating and initializing memory for a new object of a given type.
+The obvious modifications included setting the reference counter initially to zero and creating a number of free functions to free each type of object.
+\par
+In the free\emph{\_type} functions it was necessary to add further calls to decrement the reference counter of any objects referenced by the type.
+For a list list we iterate the values contained in the list, decrementing their reference (which can result in freeing that object).
+For a dictionary we decrement the reference counter on keys as well as values contained in the dictionary, since both can be objects.
+\subsection{Interesting Problems}
+
 \subsection{Assembly Output}
-
+We modified the x86 assembly to include calls to set up our memory tracking system and print statistics at the end of the program.
+\section{Results}
+\subsection{Execution}
+\subsection{Memory}
+\section{Future work}
+\section{Related Work}
 \appendix

### 2011-12-12 paper/paper.tex
replace code injection with ast transformation

@@
 This paper will focus on automatic garbage collection within a subset of the Python language, to enable a Python-like program the ability to cleanly interact with Assembly or C routines without extra intervention from the software developer.
-This paper will focus on reference counting garbage collection at the compiler level, to enable code injection which automatically frees non-referenced objects at the end of their useful lifetime.
+This paper will focus on reference counting garbage collection at the compiler level, to enable automatically freeing no-longer-referenced objects at the end of their useful lifetime.
 We will also outline the analysis steps required in the compiler which enable reference tracking, how the abstract syntax tree of the program is altered and what trade-offs are made with such altercations.
@@
 Refer to Figure~\ref{fig-comparch} for the compiler data flow.
-Each stage operates on an abstract syntax tree, the relevant new stages are the garbage collection flattening stage and the reference count code injection stage.
+Each stage operates on an abstract syntax tree, the relevant new stages are the garbage collection flattening stage and the reference count AST transformation stage.
 
@@
 The first stage is a modified flattening stage, which produces a flattened AST to aid in the reference determination.
-The next stage is the actual reference determination and code injection, where the compiler determines what variables will be objects and when assignments happen and adds appropriate calls to increment and decrement the reference counter of the object as needed.
+The next stage is the actual reference determination and AST transformation, where the compiler determines what variables will be objects and when assignments happen and adds appropriate calls to increment and decrement the reference counter of the object as needed.
 \par
@@
 \par
-The next new stage is the reference counting code injection.
+The next new stage is the reference counting AST transformation.
 In this stage we take the flattened tree and look for assignments which are setting list elements, dictionary values, class attributes, object attributes or variables.

### 2011-12-12 paper/paper.tex
talk about list addition problems and our change to program syntax by forcing an assignment

@@
 \subsection{Interesting Problems}
-
+We encountered a number of interesting questions during this implementation and summarize them below.
+\subsubsection{Addition of Lists}
+In the current runtime an entirely new list is created when two lists are added together.
+At the assignment level our compiler can determine (due to the flattened structure) that the left-hand-side and the right-hand-side of the addition need their reference counters manipulated, however the assignment from an addition to a list creates an entirely new list which should properly increment the reference counter to all objects in both lists.
+\subsubsection{Variable Assignment}
+We wanted a simple way to determine when to decrement an objects reference counter.
+Our first implementation was to add a call to decrement the counter before every assignment.
+However, because a variable can be created at any point in program execution we couldn't just blindly decrement the counter of an object since the variable will be uninitialized, and could be pointing at memory that makes it look like an object or accidentally references an object.
+\par
+The solution we took makes a slight modification to program behaviour, but we feel it is an acceptable trade-off.
+In every scope, we add an immediate assignment as the first instructions.
+This assignment is to a primitive, so it costs us no memory and then a decrement on the reference counter amounts to a function call with a no-op.
+This function call overhead could potentially be avoided by adding code to check if the variable points to an object or a primitive and is a future enhancement for performance.
 \subsection{Assembly Output}

### 2011-12-12 paper/paper.tex
Create sub sections to visually break up long-blocks-of-fail and add a new section on temp vars

@@
 \subsection{Syntax Tree Modifications}
+\subsubsection{New compiler stages}
 We added two new stages to our reference compiler.
@@
 At this stage we know every expression is operating on types.
-\par
+\subsubsection{Analysis on when to decrement an object reference}
 The next new stage is the reference counting AST transformation.
@@
 We decided on this approach to simplify determining when an object is first assigned and this allows us to simply decrement the reference counter to the object before an assign, regardless of the variables state.
+\subsubsection{Temporary Variables}
+The compiler creates temporary variables in a number of situations (namely function calls and as a result of flattened expressions).
+These temporary variables need not hang on to their memory reference past the next assignment to the actual result variable.
 \subsection{Runtime Modifications}
@@
 This function call overhead could potentially be avoided by adding code to check if the variable points to an object or a primitive and is a future enhancement for performance.
+\par
+An additional problem is lack of static single assignment.
+When a variable is re-assigned to its own name, the code should decrement the reference counter on the previous version of the object and increment it on the new version.
 \subsection{Assembly Output}

### 2011-12-12 compiler/gcrefcount.py
- Recurse on the while test stmt, and body stmt AST nodes.

@@ -175,7 +175,9 @@ class GCRefCount:
         return node
 
     def visit_While(self, node, *args, **kwargs):
-        return [node]
+        body = self.visit(node.body)
+        test = self.visit(node.test[1])
+        return [While((node.test[0],test), body, [], node.lineno)]
 
     def visit_If(self, node, *args, **kwargs):
         # recurse on the "then" and "else" statements

### 2011-12-12 compiler/test/while1.py
- Add "while" test case

@@ -1,0 +1,6 @@
+y=1
+x=[1]
+while y != 1000:
+        x=[x[0]+1]
+        y = y + 1
+print x

### 2011-12-12 compiler/gcrefcount.py
- In visit_Lambda, call getLocals() on node.code (instead of node)   This fixes the issue where variables are not properly initialized   at the beginning of a function, and decref'd at the end of the function. - Add more test cases: cycle, funcparam and funcret - Add quickcompile.sh script for testing - Delete gmon.out and profile.out during make clean target

@@ -204,9 +204,11 @@ class GCRefCount:
             raise Exception('Expected Stmt object for Lambda.code')
         decrefstmts = []
         initialassigns = []
-        for localvar in getLocals(node):
+        localAssigns = getLocals(node.code)
+        self.log.info('visit_Lambda: localAssigns=%s' % localAssigns)
+        for localvar in localAssigns:
             initialassigns.append(Assign([AssName(Name(localvar), 'OP_ASSIGN')],Const(0)))
-        for localvar in getLocals(node):
+        for localvar in localAssigns:
             decrefstmts.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(localvar)])))
         code = self.visit(node.code)
         code.nodes = initialassigns + code.nodes + decrefstmts

### 2011-12-12 compiler/pymem.c

@@ -83,10 +83,11 @@ void pymem_shutdown()
     struct timeval tv;
     node_t *ptr;
     node_t *next;
-    int i;
+    int i, leak;
 
     /* free our linked list of facts, and free any user memory that
      * failed to be free()'d and emit a warning */
+    leak = 0;
     if (head_ptr != NULL) {
         ptr = head_ptr;
         i = 0;
@@ -95,12 +96,15 @@ void pymem_shutdown()
             if (!ptr->freed) {
                 fprintf (output_fd, "WARNING: memory leak detected for allocation %d at %p\n", i, ptr->loc);
                 free(ptr->loc);
+                leak = 1;
             }
             next = ptr->next;
             free(ptr);
             ptr = next;
         }
     }
+    if (leak)
+        printf("Leak detected. See .pymem for details.\n");
 
     head_ptr = NULL;
     cur_ptr = NULL;

### 2011-12-12 compiler/quickcompile.sh

@@ -1,0 +1,14 @@
+#!/bin/sh
+# Quick script to compile a given file with the necessary options
+
+if [ "$#" -ne "1" ] ; then
+  echo "Usage: `basename $0` <source-file>"
+  exit 1
+fi
+
+PYFILE=$1
+BASENAME=`echo "$PYFILE" | cut -f1 -d.`
+SFILE="${BASENAME}.s"
+
+./compile.py $PYFILE
+gcc -o "$BASENAME" "$SFILE" -lm *.c -pg -m32

### 2011-12-12 compiler/test/cycle.py

@@ -1,0 +1,5 @@
+x = [0]
+y = [x]
+x[0] = y
+print x
+print y

### 2011-12-12 compiler/test/funcparam.py

@@ -1,0 +1,4 @@
+def f(x):
+    y = x
+    print y
+f([2])

### 2011-12-12 compiler/test/funcret.py

@@ -1,0 +1,4 @@
+def f(x):
+    y = x
+    return y
+print f([2])

### 2011-12-12 compiler/gcflattener.py
- Handle Returns as follows:   * in gcflattener.py, if we are in a Return AST node, check the     name of the variable that we are returning.  If this variable     is a temporary variable, then prevent it from being deallocated     by removing it from tmplist, thereby preventing it from getting     assigned the value zero (which is done by the call to cleanup_tempvars).     This ensures that the reference count for the object returned     from a function will be non-zero.   * in gcrefcount.py, modify visit_Lambda to keep track of all the local     assignments for a given lambda node on a stack

@@ -206,6 +206,13 @@ class GCFlattener:
     
     def visit_Return(self, node, *args, **kwargs):
         retvar, tmplist, retstmtlist = self.visit(node.value)
+        # if the return variable is in our temp list, remove it, since
+        # we need to keep the reference count at 1
+        self.log.info('visit_Return: retvar=%s' % retvar)
+        self.log.info('visit_Return: tmplist=%s' % tmplist)
+        if isinstance(retvar,Name) and retvar.name in tmplist:
+            self.log.info('visit_Return: Removing %s from tmplist' % retvar.name)
+            tmplist.remove(retvar.name)
         return retstmtlist + cleanup_tempvars(tmplist) + [Return(retvar)]
     
     def visit_CallFuncIndirect(self, node, *args, **kwargs):

### 2011-12-12 compiler/gcrefcount.py

@@ -66,6 +66,7 @@ class GCRefCount:
         self.varalloc = varalloc
         self.varset = set()
         self.log = logging.getLogger('compiler.gcrefcounter')
+        self.lambda_local_assigns = []
     
     def transform(self, node, *args, **kwargs):
         self.log.info('Starting gcrefcounter')
@@ -116,7 +117,8 @@ class GCRefCount:
                 stmtlist.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(assnode.name)])))
             self.varset.add(assnode.name)
             stmtlist.append(Assign([AssName(assnode.name,'OP_ASSIGN')], self.visit(node.expr)))
-            stmtlist.append(Discard(CallFunc(Name('inc_ref_ctr'),[Name(assnode.name)])))
+            if not isinstance(node.expr,CallFuncIndirect):
+                stmtlist.append(Discard(CallFunc(Name('inc_ref_ctr'),[Name(assnode.name)])))
             return stmtlist
         else:
             raise Exception('Unknown assignment node: %s' % assnode)
@@ -169,7 +171,14 @@ class GCRefCount:
         return node
     
     def visit_Return(self, node, *args, **kwargs):
-        return [node]
+        localAssigns = self.lambda_local_assigns[-1]
+        self.log.debug('visit_Return: localAssigns=%s' % localAssigns)
+        self.log.debug('visit_Return: node=%s' % node)
+        decrefstmts=[]
+        for localvar in localAssigns:
+            if not isinstance(node.value, Name) or node.value.name not in localAssigns:
+                decrefstmts.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(localvar)])))
+        return decrefstmts + [node]
 
     def visit_CallFuncIndirect(self, node, *args, **kwargs):
         return node
@@ -210,7 +219,9 @@ class GCRefCount:
             initialassigns.append(Assign([AssName(Name(localvar), 'OP_ASSIGN')],Const(0)))
         for localvar in localAssigns:
             decrefstmts.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(localvar)])))
+        self.lambda_local_assigns.append(localAssigns)
         code = self.visit(node.code)
+        self.lambda_local_assigns.pop()
         code.nodes = initialassigns + code.nodes + decrefstmts
         for x in node.argnames:
             self.varalloc.add_var(x)

### 2011-12-12 compiler/p3closureconvert.py

@@ -45,7 +45,7 @@ if __name__ == "__main__":
         unique = P3UniquifyVars()
         gcflatten = GCFlattener(varalloc)
         gcrefcount = GCRefCount(varalloc)
-        explicator = P3Explicate(varalloc)
+        explicator = P3Explicate(varalloc, handleLambdas=False)
         heap = P3Heapify(explicator)
         closure = P3ClosureConversion(explicator, varalloc)
 

### 2011-12-12 compiler/p3insselector.py

@@ -29,8 +29,11 @@ if __name__ == "__main__":
     import logging.config
     
     from p3declassify import P3Declassify
-    from p3explicate import P3Explicate
+    from p3wrapper import P3Wrapper
     from p3uniquifyvars import P3UniquifyVars
+    from gcflattener import GCFlattener
+    from gcrefcount import GCRefCount
+    from p3explicate import P3Explicate
     from p3heapify import P3Heapify
     from p3closureconvert import P3ClosureConversion
     from p3freevars import P3FreeVars
@@ -44,7 +47,10 @@ if __name__ == "__main__":
         varalloc = VariableAllocator()
         declassify = P3Declassify(varalloc)
         unique = P3UniquifyVars()
-        explicator = P3Explicate(varalloc)
+        wrapper = P3Wrapper()
+        gcflatten = GCFlattener(varalloc, True)
+        gcrefcount = GCRefCount(varalloc)
+        explicator = P3Explicate(varalloc, handleLambdas=False)
         heap = P3Heapify(explicator)
         closure = P3ClosureConversion(explicator, varalloc)
         flatten = P3Flattener(varalloc)
@@ -52,7 +58,10 @@ if __name__ == "__main__":
 
         ast = compiler.parseFile(testcase)
         ast = declassify.transform(ast)
+        ast = wrapper.transform(ast)
         ast = unique.transform(ast)        
+        ast = gcflatten.transform(ast)        
+        ast = gcrefcount.transform(ast)        
         ast = explicator.explicate(ast)
         ast = heap.transform(ast)
         astlist = closure.transform(ast)

### 2011-12-12 compiler/test/funcret2.py

@@ -1,0 +1,3 @@
+def f(x,y):
+    return x + y
+print f([1,2,3],[4,5,6])

### 2011-12-13 paper/paper.tex
changing to pdf instead of png
Add some valgrind results

@@
 \usepackage[pdftex]{graphicx}
+\usepackage{listings}
 
@@
 \begin{center}
-\includegraphics[scale=0.48]{compiler_flow.png}
+\includegraphics[scale=0.48]{compiler_flow.pdf}
 \end{center}
@@
 These temporary variables need not hang on to their memory reference past the next assignment to the actual result variable.
+Any temporary variable is dereferenced immediately at the end of the current scope if it is not the variable being returned.
+If it is the return variable we ignore decrementing the reference on this variable, thereby assuring the caller that the object returned is correctly reference counted for its assignment.
+If we had decremented the reference counter, due to the immediate nature of garbage collection, we would have been returning a value which did not point to an allocated memory region.
 \subsection{Runtime Modifications}
@@
 In every scope, we add an immediate assignment as the first instructions.
-This assignment is to a primitive, so it costs us no memory and then a decrement on the reference counter amounts to a function call with a no-op.
+This assignment is to a primitive, so the cost is minimal and then a decrement on the reference counter amounts to a function call with a no-op.
 This function call overhead could potentially be avoided by adding code to check if the variable points to an object or a primitive and is a future enhancement for performance.
+A program which used uninitialized (or declared) variables for program flow control would then have its runtime semantics changed.
+We felt this was a sufficient trade off since such use of an exception for program flow control is frowned upon in practice and the benefit to our design was an obvious positive.
 \par
-An additional problem is lack of static single assignment.
-When a variable is re-assigned to its own name, the code should decrement the reference counter on the previous version of the object and increment it on the new version.
+
 \subsection{Assembly Output}
 We modified the x86 assembly to include calls to set up our memory tracking system and print statistics at the end of the program.
+Running analysis tools such as valgrind on our programs before and after shows the obvious gain of going from every memory allocation leaks to zero leaks.
 \section{Results}
@@
 \subsection{Memory}
+\subsubsection{Valgrind results}
+For the trivial program
+\begin{verbatim}
+i = 0
+list = []
+
+while i !=  10000:
+    list = list + [[i]]
+    i = i + 1
+
+\end{verbatim}
+Without any garbage collection, valgrind reports 
+\begin{verbatim}
+==23467== LEAK SUMMARY:
+==23467==    definitely lost: 560,028 bytes in 20,001 blocks
+==23467==    indirectly lost: 199,004,992 bytes in 20,191 blocks
+==23467==      possibly lost: 1,375,008 bytes in 19,810 blocks
+\end{verbatim}
+With our reference counting for the same program
+\begin{verbatim}
+==23831== LEAK SUMMARY:
+==23831==    definitely lost: 0 bytes in 0 blocks
+==23831==    indirectly lost: 0 bytes in 0 blocks
+==23831==      possibly lost: 0 bytes in 0 blocks
+\end{verbatim}
 \section{Future work}

### 2011-12-13 paper/paper.tex
update a reference and some text

@@
 \category{D.3.3}{Programming Languages}{Language Constructs And Features-Dynamic Storage Management}
-\category{D.3.4}{Programming Languages}{Processors-Memory Management (garbage collection)}
+\category{D.3.4}{Programming Languages}{Processors- Memory Management (garbage collection)}
 \category{D.4.2}{Operating Systems}{Storage Management-Garbage collection}
@@
 \par
-Real-time systems are rarely implemented in languages which are garbage collected, but reference-counting garbage collection has been adapted to hard real-time systems \cite{?}.
+Real-time systems are rarely implemented in languages which are garbage collected, but reference-counting garbage collection has been adapted to hard real-time systems \cite{ritzaou}.
 Due to immediate knowledge that an object is no longer in use reference counting garbage collection can allow for destructors or finalizers to be run immediately which can enhance the clean up of system resources.
@@
 The solution we took makes a slight modification to program behaviour, but we feel it is an acceptable trade-off.
-In every scope, we add an immediate assignment as the first instructions.
+In every scope, the compiler adds an immediate assignment as the first instructions.
 This assignment is to a primitive, so the cost is minimal and then a decrement on the reference counter amounts to a function call with a no-op.
@@
 We felt this was a sufficient trade off since such use of an exception for program flow control is frowned upon in practice and the benefit to our design was an obvious positive.
-\par
 
@@
 
+\bibitem{ritzau}
+Tobias Ritzau. 1999.  Real-Time Reference Counting in RT-Java
+
 \end{thebibliography}

### 2011-12-13 compiler/gcrefcount.py
- Since CallFuncIndirect can result in a call to the runtime "create_object",   fix create_object so that it matches our new invariant that all function   calls return an object with a non-zero reference count. - Fix issues in gcrefcount.py where AssName contained a Name() wrapper when   it shouldn't. - Always call dec_ref_ctr, regardless of whether this is the initial assignment   or not.  We assign all local variables a value of zero, so it should be   safe to always call dec_ref_ctr for every assignment. - Do not initialize local variables that correspond to arguments in a Lambda.   The arg

@@ -88,7 +88,7 @@ class GCRefCount:
         decrefstmts = []
         initialassigns = []
         for localvar in getLocals(node):
-            initialassigns.append(Assign([AssName(Name(localvar), 'OP_ASSIGN')],Const(0)))
+            initialassigns.append(Assign([AssName(localvar, 'OP_ASSIGN')],Const(0)))
         for localvar in getLocals(node):
             decrefstmts.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(localvar)])))
         stmt = self.visit(node.node)
@@ -113,9 +113,9 @@ class GCRefCount:
             return [Assign([assnode], self.visit(node.expr))]
         elif isinstance(assnode, AssName):
             stmtlist = []
-            if assnode.name in self.varset:
-                stmtlist.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(assnode.name)])))
+            self.log.info('varset=%s' % self.varset)
             self.varset.add(assnode.name)
+            stmtlist.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(assnode.name)])))
             stmtlist.append(Assign([AssName(assnode.name,'OP_ASSIGN')], self.visit(node.expr)))
             if not isinstance(node.expr,CallFuncIndirect):
                 stmtlist.append(Discard(CallFunc(Name('inc_ref_ctr'),[Name(assnode.name)])))
@@ -215,8 +215,12 @@ class GCRefCount:
         initialassigns = []
         localAssigns = getLocals(node.code)
         self.log.info('visit_Lambda: localAssigns=%s' % localAssigns)
+        # Assign every local variable the value zero, except for arguments to the function
+        # This eliminates the need to know where a variable is first assigned, which may
+        # not be able to be determined statically.  
         for localvar in localAssigns:
-            initialassigns.append(Assign([AssName(Name(localvar), 'OP_ASSIGN')],Const(0)))
+            if localvar not in node.argnames:
+                initialassigns.append(Assign([AssName(localvar, 'OP_ASSIGN')],Const(0)))
         for localvar in localAssigns:
             decrefstmts.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(localvar)])))
         self.lambda_local_assigns.append(localAssigns)

### 2011-12-13 compiler/p3stackallocator.py

@@ -20,7 +20,10 @@ if __name__ == "__main__":
     import logging, logging.config
     from comp_util import *
     from p3declassify import P3Declassify
+    from p3wrapper import P3Wrapper
     from p3uniquifyvars import P3UniquifyVars
+    from gcflattener import GCFlattener
+    from gcrefcount import GCRefCount
     from p3explicate import P3Explicate
     from p3heapify import P3Heapify
     from p3closureconvert import P3ClosureConversion
@@ -35,7 +38,10 @@ if __name__ == "__main__":
         varalloc = VariableAllocator()
         declassify = P3Declassify(varalloc)
         unique = P3UniquifyVars()
-        explicator = P3Explicate(varalloc)
+        wrapper = P3Wrapper()
+        gcflatten = GCFlattener(varalloc)
+        gcrefcount = GCRefCount(varalloc)
+        explicator = P3Explicate(varalloc, handleLambdas=False)
         heap = P3Heapify(explicator)
         closure = P3ClosureConversion(explicator, varalloc)
         flatten = P3Flattener(varalloc)
@@ -43,10 +49,13 @@ if __name__ == "__main__":
 
         ast = compiler.parseFile(testcase)
         ast = declassify.transform(ast)
-        unique = unique.transform(ast)        
-        explicated = explicator.explicate(unique)
-        heaped = heap.transform(explicated)
-        astlist = closure.transform(heaped)
+        ast = wrapper.transform(ast)
+        ast = unique.transform(ast)        
+        ast = gcflatten.transform(ast)
+        ast = gcrefcount.transform(ast)         
+        ast = explicator.explicate(ast)
+        ast = heap.transform(ast)
+        astlist = closure.transform(ast)
         for ast in astlist:
             ast = flatten.flatten(ast)
             program = insselector.transform(ast)

### 2011-12-13 compiler/runtime.c

@@ -3,6 +3,8 @@
 #include <assert.h>
 #include <ctype.h>
 #include <string.h>
+#include <time.h>
+#include <sys/time.h>
 
 #include "runtime.h"
 
@@ -243,8 +245,12 @@ static big_pyobj* list_to_big(list l) {
 
 big_pyobj* create_list(pyobj length) {
     list l;
+    int i;
     l.len = project_int(length); /* this should be checked */
     l.data = (pyobj*)pymem_new(LIST, sizeof(pyobj) * l.len);
+    // Make sure list is initialized with ints, so decrefs don't fail
+    for (i=0; i < l.len; i++)
+        l.data[i]=inject_int(0);
     return list_to_big(l);
 }
 
@@ -730,10 +736,12 @@ static pyobj subscript_assign(big_pyobj* c, pyobj key, pyobj val)
     switch (c->tag) {
     case LIST:
         // XXX: inc ref count (for val only)
+	dec_ref_ctr(get_subscript(inject_big(c), key));
         inc_ref_ctr(val);
         return *list_subscript(c->u.l, key) = val;
     case DICT:
         // XXX: inc ref count (for key and val)
+	dec_ref_ctr(get_subscript(inject_big(c), key));
         inc_ref_ctr(key);
         inc_ref_ctr(val);
         return *dict_subscript(c->u.d, key) = val;
@@ -980,6 +988,7 @@ big_pyobj* create_object(pyobj cl) {
         exit(-1);
     }
     ret->u.obj.attrs = create_hashtable(2, attrname_hash, attrname_equal, OBJECT);
+    inc_ref_ctr(inject_big(ret));
     return ret;
 }
 
@@ -1369,7 +1378,15 @@ void inc_ref_ctr(pyobj v) {
 }
 
 void dec_ref_ctr(pyobj v) {
+    struct timeval start, end, result;
+    int ret;
+
     if (is_big(v)) {
+	//printf("decref(%p): ", v);
+	//print_any(v);
+	//printf("\n");
+        ret = gettimeofday(&start, NULL);
+        assert (ret > -1);
         big_pyobj *val = project_big(v);
         //printf("decrementing ref_ctr at %d addr %d on ", val, val->ref_ctr);
         //print_any(v);
@@ -1382,6 +1399,35 @@ void dec_ref_ctr(pyobj v) {
         if ( val->ref_ctr == 0 ) {
             release(val);
         }
-
+        ret = gettimeofday(&end, NULL);
[…]

### 2011-12-13 compiler/test/dict_subscript.py

@@ -1,0 +1,3 @@
+a={}
+a[0] = 5
+print a[0]

### 2011-12-13 compiler/test/list_subscript.py

@@ -1,0 +1,3 @@
+a = [[]]
+a[0] = 1
+

### 2011-12-13 compiler/test/simpleclass.py

@@ -1,0 +1,3 @@
+class A:
+    0
+a = A()

### 2011-12-13 compiler/test/simplelist.py

@@ -1,0 +1,5 @@
+a=[0,1,2,3]
+a[0]=[0,1,2,3]
+a[1]=[0,1,2,3]
+a[2]=[0,1,2,3]
+a[3]=[0,1,2,3]

### 2011-12-13 compiler/test/while2.py

@@ -1,0 +1,6 @@
+i = 0
+lst=[0]
+while i != 10000: 
+    b = [lst[0]]
+    lst[0] = b 
+    i = i + 1

### 2011-12-13 compiler/runtime.c
- Add dec_ref_ctr_rec function, and make dec_ref_ctr delegate to this   function.  Now dec_ref_ctr calculates the latency associated with   a de-allocation.

@@ -19,6 +19,7 @@ static void print_list(pyobj pyobj_list);
 static void print_dict(pyobj dict);
 static list list_add(list x, list y);
 static void print_pyobj(pyobj x);
+static void dec_ref_ctr_rec(pyobj v);
 
 int tag(pyobj val) {
     return val & MASK;
@@ -736,12 +737,12 @@ static pyobj subscript_assign(big_pyobj* c, pyobj key, pyobj val)
     switch (c->tag) {
     case LIST:
         // XXX: inc ref count (for val only)
-	dec_ref_ctr(get_subscript(inject_big(c), key));
+	dec_ref_ctr_rec(get_subscript(inject_big(c), key));
         inc_ref_ctr(val);
         return *list_subscript(c->u.l, key) = val;
     case DICT:
         // XXX: inc ref count (for key and val)
-	dec_ref_ctr(get_subscript(inject_big(c), key));
+	dec_ref_ctr_rec(get_subscript(inject_big(c), key));
         inc_ref_ctr(key);
         inc_ref_ctr(val);
         return *dict_subscript(c->u.d, key) = val;
@@ -1225,10 +1226,10 @@ void iterate_and_release_table(struct hashtable* h, int dec_keys) {
         pyobj v = *(pyobj *)hashtable_iterator_value(itr);
 
         if ( dec_keys ) {
-            dec_ref_ctr(k);
+            dec_ref_ctr_rec(k);
         }
 
-        dec_ref_ctr(v);
+        dec_ref_ctr_rec(v);
     } while (hashtable_iterator_advance(itr));
     // none of the above functions which use iterators clean this up...
     free(itr);
@@ -1253,7 +1254,7 @@ void free_class(big_pyobj* o) {
      */
     unsigned int i=0;
     // decrement reference count for list of parent classes
-    dec_ref_ctr(o->u.cl.parents);
+    dec_ref_ctr_rec(o->u.cl.parents);
     // decrement reference counts for all values in the attribute hash table  
     // (but not keys since the keys are just strings that are not managed
     //  by the runtime)
@@ -1277,7 +1278,7 @@ void free_class(big_pyobj* o) {
  */
 void free_object(big_pyobj* o) {
     // decrement the class reference
-    dec_ref_ctr(o->u.obj.clazz);
[…]

### 2011-12-13 compiler/runtime.c
- Commented out latency calculation for now...

@@ -1397,13 +1397,13 @@ void dec_ref_ctr(pyobj v) {
     struct timeval start, end, result;
     int ret;
 
-    ret = gettimeofday(&start, NULL);
-    assert (ret > -1);
+    //ret = gettimeofday(&start, NULL);
+    //assert (ret > -1);
     dec_ref_ctr_rec(v);
-    ret = gettimeofday(&end, NULL);
-    assert (ret > -1);
-    timeval_subtract(&result, &end, &start);
-    fprintf(stderr, "dec_ref_ctr took %ld.%ld seconds\n", result.tv_sec, result.tv_usec);
+    //ret = gettimeofday(&end, NULL);
+    //assert (ret > -1);
+    //timeval_subtract(&result, &end, &start);
+    //fprintf(stderr, "dec_ref_ctr took %ld.%ld seconds\n", result.tv_sec, result.tv_usec);
 }
 
 /* Subtract the `struct timeval' values X and Y,

### 2011-12-13 paper/paper.tex
- Modified the intro.

@@
 \begin{abstract}
-Integrating a modern dynamic language in to existing systems-level languages (Assembly or C) easily has a number of challenges.
-This paper will focus on automatic garbage collection within a subset of the Python language, to enable a Python-like program the ability to cleanly interact with Assembly or C routines without extra intervention from the software developer.
-This paper will focus on reference counting garbage collection at the compiler level, to enable automatically freeing no-longer-referenced objects at the end of their useful lifetime.
-We will also outline the analysis steps required in the compiler which enable reference tracking, how the abstract syntax tree of the program is altered and what trade-offs are made with such altercations.
+Garbage collection in high level languages has enabled increased programmer productivity, but at the cost of performance and ease of implementation for the language designer.
+This paper will explore these trade-offs through an implementation of automatic garbage collection for a subset of the Python language.
+Our implementation will extend an existing Python to x86 compiler to use reference counting garbage collection.
+Reference counting is chosen for its ease of implementation, and because it is less invasive when one of the design goals is to integrate with assembly or C routines without extra intervention from the software developer.
+We outline the changes to the existing compiler to enable reference counting, how the abstract syntax tree of the program is altered and what trade-offs are made with our design choices.
 \end{abstract}

### 2011-12-13 paper/paper.tex
More changes to introduction (last commit was actually a change to the abstract).

@@
 
-Python is a modern high level language which has a reference counting garbage collector at its core.
-It is also, by default, an interpreted language.
-In many cases the usefulness of Python as a productive language is highly desired and a trade-off is made in terms of performance.
-While numerous optimizations have been made to the Python run time there is still a usefulness to compile a high level language directly to machine code.
-Most high level languages do not allow for direct memory management, instead leaving allocation and de-allocation of memory to the interpreter and garbage collection process.
-There are two main classifications for garbage collectors: generational or tracing collectors and reference-counting collectors.
-Generational or tracing collectors make periodic inspection of all objects to determine if they are reachable by a pointer or object reference.
-These are the modern batch of garbage collectors and have the best performance
-Given a naive implementation of a direct Python to x86-assembly compiler which disregards all memory management, we explore the choices and consequences of adding a reference counting garbage collector.
-Our base Python-like language includes classes, objects, functions, lambdas, simple math, lists, dictionaries and integer primitives.
-\subsection{Disadvantages}
+Python is a modern, high level language that uses dynamic typing and is run within an interpreter.
+It manages memory for the programmer, and is considered by many to be a good "prototyping" language, in which programmers can quickly and easily develop an application.
+Automatic memory management plays a large part in supporting this role, as it relieves the programmer from the tedious task of remembering when to de-allocate memory.
+For language designers, deciding what form of memory management to incorporate into their language can be a difficult task.
+The choice will depend on many factors, including the designer's vision on the general usefulness of the language, the environment where the language will be primary used, performance requirements, program semantics, integration with legacy systems and many other factors.
+The choice also has significant implications on the implementation and therefore warrants careful consideration.
+
+There are currently two widely used strategies to perform automatic memory management.
+One such strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of "root" objects.
+This set of objects includes the set of all globals, as well as the objects which are referencable from the current stack frame.
+Any objects that are traversed during this operation are considered reachable and therefore cannot be de-allocated.
+Conversely, any objects not traversed are no longer reachable, and can be de-allocated.
+The other primary strategy for automated memory management, is reference counting garbage collection.
+Reference counting works by keeping a count of references to a particular object.
+References are tracked by incrementing and decrementing an object's reference count when an assignment to a pointer occurs, or when a pointer variable goes out of scope.
+When the reference count reaches zero, it is safe to de-allocate the object.
+
+The performance characteristics of each strategy can vary considerably, depending on implementation.
+Tracing garbage collectors can sometimes be associated with long "pause" times, where application work must be paused to allow the entire contents of the heap to be scanned by the garbage collector.
+These long pause times have largely been solved with new algorithms that improve upon the original naive mark-and-sweep algorithm.
+It is typical that these algorithms tradeoff throughput for responsiveness.
+However, these pause times can still be an issue, especially in real time systems, where a hard guarantees on latency are needed.
+Reference counting has been discussed as a solution to this problem, but classic implementations of reference counting can have serious performance concerns when one de-allocation results in cascading de-allocations\cite{boehm}.
+
+Given a existing implementation of a Python to x86-assembly compiler which disregards all memory management, we explore the consequences of adding a reference counting garbage collector.
+Our base Python-like language includes classes, objects, functions, lambdas, addition, unary subtraction, lists, dictionaries and integer primitives.
+The remainder of the paper is organized as follows.
+Section~\ref{sec:related} discusses related work.
+Section~\ref{sec:implementation} discusses the implementation of reference counting for our compiler.
+Section~\ref{sec:results} evaluates our compiler for performance and correctness.
+Section~\ref{sec:future} discusses potential future work.
+Finally, section~\ref{sec:conclusion} concludes.
+
+\section{Related Work}
+\label{sec:related}
 In general a reference count garbage collector will have shorter pause times for the garbage collection phase, but will incur a higher performance penalty \cite{joisha}\cite{blackburn}.
@@
 Without extra work a strictly reference-counting collector will leak memory when references contain cycles, and multiple algorithms have been presented as solutions to this problem.
-\subsection{Advantages}
+\textit{Citation?}
+
 One potential advantage that reference-counting garbage collectors have is the ability to be implemented in hardware \cite{joao}.
 Given the significant cost of garbage collection \cite{hertz} any hardware assisted acceleration would be highly desirable if it is flexible enough to adapt across multiple garbage collection techniques.
-\par
+
 Real-time systems are rarely implemented in languages which are garbage collected, but reference-counting garbage collection has been adapted to hard real-time systems \cite{ritzaou}.
 Due to immediate knowledge that an object is no longer in use reference counting garbage collection can allow for destructors or finalizers to be run immediately which can enhance the clean up of system resources.
+
+\par
 \section{Implementation}
+\label{sec:implementation}
+
 We develop a base compiler which generates x86 assembly code and then alter the compiler to include reference counting and garbage collection.
+
 \subsection{Language}
@@
 \subsection{Syntax Tree Modifications}
+
 \subsubsection{New compiler stages}
@@
 At this stage we know every expression is operating on types.
+
 \subsubsection{Analysis on when to decrement an object reference}
@@
 We decided on this approach to simplify determining when an object is first assigned and this allows us to simply decrement the reference counter to the object before an assign, regardless of the variables state.
+
 \subsubsection{Temporary Variables}
@@
 If we had decremented the reference counter, due to the immediate nature of garbage collection, we would have been returning a value which did not point to an allocated memory region.
+
 \subsection{Runtime Modifications}
@@
 For a dictionary we decrement the reference counter on keys as well as values contained in the dictionary, since both can be objects.
+
 \subsection{Interesting Problems}
 We encountered a number of interesting questions during this implementation and summarize them below.
+
 \subsubsection{Addition of Lists}
@@
 At the assignment level our compiler can determine (due to the flattened structure) that the left-hand-side and the right-hand-side of the addition need their reference counters manipulated, however the assignment from an addition to a list creates an entirely new list which should properly increment the reference counter to all objects in both lists.
+
 \subsubsection{Variable Assignment}
@@
 Running analysis tools such as valgrind on our programs before and after shows the obvious gain of going from every memory allocation leaks to zero leaks.
+
 \section{Results}
+\label{sec:results}
+
 \subsection{Execution}
+
 \subsection{Memory}
+
 \subsubsection{Valgrind results}
@@
 \end{verbatim}
+
 \section{Future work}
-\section{Related Work}
+\label{sec:future}
+
+\section{Conclusions}
+\label{sec:conclusion}
+
+\subsection{Disadvantages}
+
+\subsection{Advantages}
+
 \appendix

### 2011-12-13 paper/paper.tex
- Modifications to Implementation section (runtime modifications)

@@
 Reference counting is chosen for its ease of implementation, and because it is less invasive when one of the design goals is to integrate with assembly or C routines without extra intervention from the software developer.
-We outline the changes to the existing compiler to enable reference counting, how the abstract syntax tree of the program is altered and what trade-offs are made with our design choices.
+We outline the changes to the existing compiler to enable reference counting, including how the abstract syntax tree of the program is altered, and then discuss the trade-offs made by our design choices.
 \end{abstract}
@@
 Python is a modern, high level language that uses dynamic typing and is run within an interpreter.
-It manages memory for the programmer, and is considered by many to be a good "prototyping" language, in which programmers can quickly and easily develop an application.
+It manages memory for the programmer, and is considered by many to be a good ``prototyping'' language, in which programmers can quickly and easily develop an application.
 Automatic memory management plays a large part in supporting this role, as it relieves the programmer from the tedious task of remembering when to de-allocate memory.
@@
 There are currently two widely used strategies to perform automatic memory management.
-One such strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of "root" objects.
-This set of objects includes the set of all globals, as well as the objects which are referencable from the current stack frame.
+One such strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.
+This set of objects includes all global variables, as well as the objects which are referenceable from the current stack frame.
 Any objects that are traversed during this operation are considered reachable and therefore cannot be de-allocated.
@@
 The performance characteristics of each strategy can vary considerably, depending on implementation.
-Tracing garbage collectors can sometimes be associated with long "pause" times, where application work must be paused to allow the entire contents of the heap to be scanned by the garbage collector.
+Tracing garbage collectors can sometimes be associated with long pause times, where application work must be temporarily stopped to allow the entire contents of the heap to be scanned by the garbage collector.
 These long pause times have largely been solved with new algorithms that improve upon the original naive mark-and-sweep algorithm.
@@
 
-Given a existing implementation of a Python to x86-assembly compiler which disregards all memory management, we explore the consequences of adding a reference counting garbage collector.
+Given an existing implementation of a Python to x86-assembly compiler which lacks any memory management, we explore the consequences of adding a reference counting garbage collector.
 Our base Python-like language includes classes, objects, functions, lambdas, addition, unary subtraction, lists, dictionaries and integer primitives.
@@
 
+In \cite{levanoni}, Levanoni, et. al. show that reference counting is a valid approach for multprocessor systems.
+
 \par
@@
 
-We develop a base compiler which generates x86 assembly code and then alter the compiler to include reference counting and garbage collection.
+We start with an existing compiler based upon previous work\cite{siek}.
+This compiler converts a subset of the Python language to x86 assembly and supports one of the core features of Python, namely that everything is a first class object.
+Prior to our modifications, the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs, with minimal memory constraints.
+After our modifications, the compiler properly maintains reference counts for all objects created, and automatically de-allocates objects when there are no longer any references to them.
 
-\subsection{Language}
-A subset of Python has been chosen which includes functions, objects, lambdas, and the primitives integer, boolean.
-We implement this as a reference compiler which is extended to include reference counting.
-This language has no built in parallelism but Levanoni, et.
-al.
-have shown that reference counting is a valid approach for multprocessor systems.\cite{levanoni}
-\par
-Starting from our reference compiler, we treat any object which is not an integer or boolean primitive as a C structure.
-This structure was modified to include a reference counter and in all places where such a structure was allocated we converted the allocation to use a statistics tracking allocator and set the reference count to an initial value of 0.
-\par
-We chose to implement our own memory routines primarily to investigate object lifetime (initial allocation time and final de-allocation time) as well as allow us the ability to detect memory leaks.
-Since every allocation of an object in our language uses this wrapper we know exactly how much memory has been allocated and what objects were never de-allocated.
-\par
-The basic premise of such routines is to wrap the provided malloc and free, request slightly more memory than required.
-That extra memory is what where the reference counter is stored and the pointer location is returned which skips the counter.
-The memory routines also create a linked list to track all of the allocations in a structure for statistics gathering and leak detection.
+Implementing reference counting requires a minimum set of steps to be taken.
+
+\begin{itemize}
+\item Objects must include a counter that tracks the number of references to that object.
+\item Functions to increment and decrement an objects reference count must be created.
+\item Assignment of an object to a pointer or \textit{reference} should result in correct updates to an object's reference count.
+\item Reference counts must be decremented when a program reaches the end of a given scope, since local variables are destroyed.
+\item When an object's reference count reaches zero, it is no longer reachable and can be safely destroyed.
+
+\end{itemize}
+
+These steps require modifications to the architecture of our compiler, as well as the supporting runtime library responsible for creating objects and allocating memory.
+We discuss these modifications in the following sections.
+
+\subsection{Runtime Modifications}
+Our runtime library is implemented in C so that it can be easily called by the assembly programs created by our compiler.
+The runtime library contains abstractions such as \texttt{create\_list}, \texttt{create\_dict} and \texttt{create\_object} that take care of allocating and initializing memory for a new object of the specified type.
+These functions return a pointer to the memory allocated, and the pointer is \textit{tagged} to indicate that it represents a non-primitive type.
+In our reference compiler, any non-primitive object must be a pointer to a C structure called \texttt{big\_pyobj}.
+To support reference counts, this C structure is modified to include an int field called \texttt{ref\_ctr}.
+Then all of the functions to create objects are modified to set the initial reference count to zero.
+
+To support incrementing and decrementing an object's reference count, we create corresponding functions in our runtime called \texttt{inc\_ref\_ctr} and \texttt{dec\_ref\_ctr}.
+The increment function is trivial in that it simply de-references the object pointer and increments the \texttt{ref\_ctr} field.
+However, the decrement function potentially has more work to do than just de-referencing the pointer and decrementing the reference count.
+According to the steps for reference counting mentioned previously, we need to de-allocate an object whenever its reference count reaches zero.
+To provide this behavior, the \texttt{dec\_ref\_ctr} function has to check if the reference count has reached zero, and if so, delegate to a corresponding \texttt{free\_\textit{type}} method to free an object of the given type.
+
+In the \texttt{free\_\textit{type}} functions, it is necessary to add further calls to decrement the reference counter of any objects referenced by the object being de-allocated.
+For the case of a list, we have the \texttt{free\_list} function, which iterates over each element in the list and decrements its reference count.
+For a dictionary, we have the \texttt{free\_dict} function, which decrements the reference counter on keys as well as values contained in the dictionary, since both can be objects.
+Note that for any case of an object that references other objects, this can can result in de-allocation of the referenced object, if the only reference to it is held by the parent object.
+In the worst case, this could lead to a number of de-allocations equivalent to the number of nodes in the object graph created by the parent object.
+
+Since every object in our system may contain references to another object, we modify our runtime to correctly increment the reference count when such as reference is established.
+For example, an assignment to a list or dictionary subscript results in a call to the \texttt{set\_subscript} runtime function.
+This function is modified to first decrement the reference count for any object that may have been previously referenced by the given subscript, and then increment the reference count for the new object that is assigned to the given subscript.
+Similarly, a closure maintains a reference to its list of free variables, and a class object maintains a reference to its base classes.
+In each case, we modify the appropriate runtime function to ensure that these reference counts are correctly maintained.
+
+In some cases, the type definition for an object in our runtime did not aggregate a \texttt{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \texttt{class\_struct}).
+For this reason, significant changes were made to the runtime to ensure that any type that references another type, has a generic pointer to a \texttt{big\_pyobj} structure, so that the reference counts can be correctly maintained.
+To ensure that we did not make any mistakes in this process, we create a runtime test suite that uses the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are de-allocated when their reference count reaches zero.
+This runtime test suite is important to unit test the runtime library, before attempting to use it from our compiler.
+
+Finally, in order to verify the correctness of our implementation, we create a set of abstractions on top of the system allocation functions \texttt{malloc} and \texttt{free}.
+These abstractions are part of our runtime and provide us the following benefits.
+
+\begin{enumerate}
+\item Ability to detect memory leaks
+\item Ability to track allocations by type/size
+\item Ability to track the time of allocation/de-allocation
+\end{enumerate}
+
+The abstractions are implemented in a new module appropriately called \texttt{pymem}.
+Every allocation of an object in our runtime is modified to use the \texttt{pymem\_new} and \texttt{pymem\_free} instead of \texttt{malloc} and \texttt{free}.
+As a result of this, we can determine exactly how much memory has been allocated and what objects have still not been de-allocated, by calling the \texttt{pymem\_print\_statistics} function at any given program point.
+In practice, this is only called at the end of each program to ensure that all memory has been de-allocated.
+Since we can track the time of allocation and de-allocation, this module can be beneficial for investigating object lifetime as well.
+
+The \texttt{pymem\_new} routine works by allocating slightly more memory than is requested by the user.
+This extra memory is used to store a pointer to a linked list node that contains the details of the allocation, including the type of object being created, the requested size, the pointer returned by \texttt{malloc} and the time of allocation.
+The caller is then returned the portion of memory immediately after this pointer.
+Subsequent calls to \texttt{pymem\_free} then have access to the linked list node via the pointer, and can update the node to mark the memory as de-allocated, as well as the de-allocation time.
+The memory can then be freed independently of the linked list containing the statistics.
 \begin{figure}
@@
 If we had decremented the reference counter, due to the immediate nature of garbage collection, we would have been returning a value which did not point to an allocated memory region.
-
-\subsection{Runtime Modifications}
-Our runtime consisted of C functions which took care of allocating and initializing memory for a new object of a given type.
-The obvious modifications included setting the reference counter initially to zero and creating a number of free functions to free each type of object.
-\par
-In the free\emph{\_type} functions it was necessary to add further calls to decrement the reference counter of any objects referenced by the type.
-For a list list we iterate the values contained in the list, decrementing their reference (which can result in freeing that object).
-For a dictionary we decrement the reference counter on keys as well as values contained in the dictionary, since both can be objects.
 
@@
 \label{sec:future}
+
+Optimizations.
 
@@
 
+\bibitem{siek}
+Jeremy G. Siek. 2011.  A Problem Course in Compilation: From Python to x86 Assembly
+
 \end{thebibliography}

### 2011-12-14 paper/paper.tex
- Changes up through 3.3.1 - Expressions

@@
 The memory can then be freed independently of the linked list containing the statistics.
+
+\subsection{Compiler}
+
+Figure~\ref{fig-comparch} shows the compiler data flow after modifying our compiler for reference counting.
+Each stage operates on an abstract syntax tree (AST).
+The relevant new stages are the pre-explicate flatten stage, and the reference counting AST transformation stage.
+
 \begin{figure}
@@
 
-\subsection{Compiler Flow}
-Refer to Figure~\ref{fig-comparch} for the compiler data flow.
-Each stage operates on an abstract syntax tree, the relevant new stages are the garbage collection flattening stage and the reference count AST transformation stage.
+\subsubsection{Reference Counting AST Transformation}
 
-\subsection{Syntax Tree Modifications}
+Recall that in order to properly maintain reference counts, we need to increment the reference count for an object when a reference is created for it, and decrement the reference count when a reference is destroyed.
+References are created by assigning a memory location (which contains an object) to a variable.
+References are destroyed when a variable is changed to refer to a different object, or when a variable ceases to exist when a program reaches the end of a given scope.
 
-\subsubsection{New compiler stages}
-We added two new stages to our reference compiler.
-The first stage is a modified flattening stage, which produces a flattened AST to aid in the reference determination.
-The next stage is the actual reference determination and AST transformation, where the compiler determines what variables will be objects and when assignments happen and adds appropriate calls to increment and decrement the reference counter of the object as needed.
-\par
-Our initial assumption was that we could easily add the increment and decrement reference counter operations in the reference compilers flatten stage, however because this stage is after type explication and there are variables created which do not correspond to a type in the language the pre-flatten stage was necessary.
-This stage converts all expressions in to assignments to temporary variables.
-At this stage we know every expression is operating on types.
+The most natural way to implement this in our compiler, is to transform our input programs AST such that any assignment to a variable is preceded by a decrement to the reference count for the object the variable refers to, and is succeeded by an increment to the reference count for the newly assigned object.
+This is the approach we took, but there are some subtleties with respect to where this operation is placed within the compiler.
+Our initial hypothesis was that we could easily add the increment and decrement reference counter operations in the existing ``Flatten AST'' stage.
+We quickly realized that because this stage is after type explication, there exist variables which are not \textit{tagged} with a type.
+Since we only want to decrement reference counts for \textit{tagged} types\footnote{Actually, we only want to modify reference counts for a subset of tagged types that only includes ``big'' types.
+However, to simplify our implementation, we transform our AST to always call our increment and decrement runtime functions, but these functions will ignore the primitive types of \texttt{int} and \texttt{boolean}.}, the only option is to put this transformation stage before type explication.
 
-\subsubsection{Analysis on when to decrement an object reference}
-The next new stage is the reference counting AST transformation.
-In this stage we take the flattened tree and look for assignments which are setting list elements, dictionary values, class attributes, object attributes or variables.
-The compiler determines what variables are created in the current scope and injects code at the beginning of that scope to set the value initially.
-We decided on this approach to simplify determining when an object is first assigned and this allows us to simply decrement the reference counter to the object before an assign, regardless of the variables state.
+\subsection{Challenges}
+We encountered a number of interesting challenges related to our design choices in modifying the compiler and summarize them below.
+
+\subsubsection{Expressions}
+Prior to our explicate phase, complex expressions are not flattened.
+This leads to problems where memory is leaked, since no assignments ever occur to temporary objects.
+Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between to lists to the variable \texttt{a}.
+The assignment to the variable \texttt{a} results in a corresponding increment reference count on the result of the list addition.
+However, the temporary lists, \texttt{[1]} and \texttt{[2]} are leaked if we don't take any additional steps.
+To overcome this problem, we introduce a intermediate flattening stage, prior to our reference counting AST transformation.
+This results in the corresponding code given in Listing~\ref{lst:flattened}.
+
+\begin{figure}[h!]
+\begin{lstlisting}[
+language=Python,
+frame=topline,
+caption={A simple input program that leaks memory if not flattened.},
+captionpos=b,
+label=lst:simpleleak
+]
+a = [1] + [2]
+\end{lstlisting}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{lstlisting}[
+language=Python,
+frame=topline,
+caption={The flattened version of listing \ref{lst:simpleleak}.},
+captionpos=b,
+label=lst:flattened
+]
+tmp0 = [1]
+tmp1 = [2]
+tmp2 = tmp0 + tmp1
+\end{lstlisting}
+\end{figure}
 
@@
 If we had decremented the reference counter, due to the immediate nature of garbage collection, we would have been returning a value which did not point to an allocated memory region.
-
-\subsection{Interesting Problems}
-We encountered a number of interesting questions during this implementation and summarize them below.
 
@@
 
-\appendix
-\section{Appendix Title}
-
-This is the text of the appendix, if you need one.
+%\appendix
+%\section{Appendix Title}
+% This is the text of the appendix, if you need one.
 

### 2011-12-14 paper/paper.tex
- Oops, one tiny change.

@@
 tmp2 = tmp0 + tmp1
+a = tmp2
 \end{lstlisting}

### 2011-12-14 paper/paper.tex
Through 3.3.2.

--- 2011-12-14 paper/paper.tex
@@
 However, to simplify our implementation, we transform our AST to always call our increment and decrement runtime functions, but these functions will ignore the primitive types of \texttt{int} and \texttt{boolean}.}, the only option is to put this transformation stage before type explication.
+In addition to modifying the AST prior before/after an assignment, the reference counting transformation is also responsible for determining which variables are local to a given scope and decrementing the counts for these references at the end of the given program scope.
 
@@
 a = [1] + [2]
+a = 0
+print a
 \end{lstlisting}
@@
 tmp2 = tmp0 + tmp1
+a = tmp2
+a = 0
+print a
 \end{lstlisting}
@@
 
-\subsubsection{Temporary Variables}
-The compiler creates temporary variables in a number of situations (namely function calls and as a result of flattened expressions).
-These temporary variables need not hang on to their memory reference past the next assignment to the actual result variable.
-Any temporary variable is dereferenced immediately at the end of the current scope if it is not the variable being returned.
-If it is the return variable we ignore decrementing the reference on this variable, thereby assuring the caller that the object returned is correctly reference counted for its assignment.
-If we had decremented the reference counter, due to the immediate nature of garbage collection, we would have been returning a value which did not point to an allocated memory region.
+\subsubsection{Temporary variables}
+As a result of introducing the intermediate flattening stage, we introduce additional references to an object which hang around until the end of the given program scope.
+This is undesirable because it changes the de-allocation behavior of the program from the expected behavior.
+For original input program in figure~\ref{lst:simplebreak}, the expectation is that when the variable \texttt{a} is assigned the value zero, the reference count for the list object should be decremented to zero and the object de-allocated.
+However, in the flattened program, this is no longer the case since the \texttt{tmp2} variable continues to reference the object.
+The de-allocation is therefore delayed until the end of the current scope.
+To handle this situation, we keep track of all temporary variables introduced as a result of flattening an expression.
+These temporary variables are then assigned a value of zero immediately after the statement containing the expression in which they were introduced.
+This will lead to an immediate decrement of the reference count after the reference counting AST transformation.
+Figure~\ref{lst:fixflattened} shows the result of the flattening transformation after this change is made.
 
-\subsubsection{Addition of Lists}
-In the current runtime an entirely new list is created when two lists are added together.
-At the assignment level our compiler can determine (due to the flattened structure) that the left-hand-side and the right-hand-side of the addition need their reference counters manipulated, however the assignment from an addition to a list creates an entirely new list which should properly increment the reference counter to all objects in both lists.
+\begin{lstlisting}[
+language=Python,
+frame=topline,
+caption={The flattened version of listing \ref{lst:simpleleak} with immediate de-allocation of temporary variables.},
+captionpos=b,
+label=lst:flattened
+]
+tmp0 = [1]
+tmp1 = [2]
+tmp2 = tmp0 + tmp1
+a = tmp2
+tmp2 = 0
+tmp1 = 0
+tmp0 = 0
+a = 0
+print a
+\end{lstlisting}
+\end{figure}
 
@@
 We felt this was a sufficient trade off since such use of an exception for program flow control is frowned upon in practice and the benefit to our design was an obvious positive.
+
+\subsubsection{Addition of Lists}
+In the current runtime an entirely new list is created when two lists are added together.
+At the assignment level our compiler can determine (due to the flattened structure) that the left-hand-side and the right-hand-side of the addition need their reference counters manipulated, however the assignment from an addition to a list creates an entirely new list which should properly increment the reference counter to all objects in both lists.
+
+\subsubsection{Functions}
+
+The compiler creates temporary variables in a number of situations (namely function calls and as a result of flattened expressions).
+These temporary variables need not hang on to their memory reference past the next assignment to the actual result variable.
+Any temporary variable is dereferenced immediately at the end of the current scope if it is not the variable being returned.
+If it is the return variable we ignore decrementing the reference on this variable, thereby assuring the caller that the object returned is correctly reference counted for its assignment.
+If we had decremented the reference counter, due to the immediate nature of garbage collection, we would have been returning a value which did not point to an allocated memory region.
 

### 2011-12-14 paper/paper.tex
fix label issue I introduced

@@
 This is undesirable because it changes the de-allocation behavior of the program from the expected behavior.
-For original input program in figure~\ref{lst:simplebreak}, the expectation is that when the variable \texttt{a} is assigned the value zero, the reference count for the list object should be decremented to zero and the object de-allocated.
+For original input program in figure~\ref{lst:simpleleak}, the expectation is that when the variable \texttt{a} is assigned the value zero, the reference count for the list object should be decremented to zero and the object de-allocated.
 However, in the flattened program, this is no longer the case since the \texttt{tmp2} variable continues to reference the object.
@@
 
+\begin{figure}[h!]
 \begin{lstlisting}[
@@
 captionpos=b,
-label=lst:flattened
+label=lst:fixflattened
 ]

### 2011-12-14 paper/paper.tex
update references

@@
 Reference counting is chosen for its ease of implementation, and because it is less invasive when one of the design goals is to integrate with assembly or C routines without extra intervention from the software developer.
-We outline the changes to the existing compiler to enable reference counting, including how the abstract syntax tree of the program is altered, and then discuss the trade-offs made by our design choices.
+We outline the changes to the existing compiler to enable reference counting, including how the abstract syntax tree of the program is altered, and then discuss the implications of our design choices.
 \end{abstract}
@@
 One such strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.
-This set of objects includes all global variables, as well as the objects which are referenceable from the current stack frame.
+This set of objects includes all global variables, as well as the objects which are reachable from the current stack frame.
 Any objects that are traversed during this operation are considered reachable and therefore cannot be de-allocated.
@@
 These long pause times have largely been solved with new algorithms that improve upon the original naive mark-and-sweep algorithm.
-It is typical that these algorithms tradeoff throughput for responsiveness.
+It is typical that these algorithms trade-off throughput for responsiveness.
 However, these pause times can still be an issue, especially in real time systems, where a hard guarantees on latency are needed.
@@
 Without extra work a strictly reference-counting collector will leak memory when references contain cycles, and multiple algorithms have been presented as solutions to this problem.
-\textit{Citation?}
+\cite{lin}\cite{trancon}
 
@@
 \softraggedright
+
+\bibitem{boehm}
+Hans-J.
+Boehm.
+2004.
+The space cost of lazy reference counting.
+In Proceedings of the 31st ACM SIGPLAN-SIGACT symposium on Principles of programming languages (POPL '04).
+ACM, New York, NY, USA, 210-219.
+DOI=10.1145/964001.964019
+
 \bibitem{joisha}
@@
 
-\bibitem{boehm}
-Hans-J.
-Boehm.
-2004.
-The space cost of lazy reference counting.
-In Proceedings of the 31st ACM SIGPLAN-SIGACT symposium on Principles of programming languages (POPL '04).
-ACM, New York, NY, USA, 210-219.
-DOI=10.1145/964001.964019
+\bibitem{lin}
+Chin-Yang Lin and Ting-Wei Hou.
+2010.
+An efficient approach to cyclic reference counting based on a coarse-grained search.
+Inf.
+Process.
+Lett.
+111, 1 (December 2010), 1-10.
+DOI=10.1016/j.ipl.2010.10.004
+
+\bibitem{trancon}
+Baltasar Trancón y Widemann.
+2008.
+A reference-counting garbage collection algorithmfor cyclical functional programming.
+In Proceedings of the 7th international symposium on Memory management (ISMM '08).
+ACM, New York, NY, USA, 71-80.
+DOI=10.1145/1375634.1375645
 
@@
 
-\bibitem{ritzau}
+\bibitem{ritzaou}
 Tobias Ritzau. 1999.  Real-Time Reference Counting in RT-Java

### 2011-12-14 paper/paper.tex
Updates to paper.

@@
 \usepackage{listings}
+
+\lstset{
+basicstyle=\ttfamily,
+numberstyle=\small,
+commentstyle=\textit
+stepnumber=1
+}
 
@@
 It manages memory for the programmer, and is considered by many to be a good ``prototyping'' language, in which programmers can quickly and easily develop an application.
-Automatic memory management plays a large part in supporting this role, as it relieves the programmer from the tedious task of remembering when to de-allocate memory.
+Automatic memory management plays a large part in supporting this role, as it relieves the programmer from the tedious task of remembering when to deallocate memory.
 For language designers, deciding what form of memory management to incorporate into their language can be a difficult task.
@@
 One such strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.
-This set of objects includes all global variables, as well as the objects which are reachable from the current stack frame.
-Any objects that are traversed during this operation are considered reachable and therefore cannot be de-allocated.
-Conversely, any objects not traversed are no longer reachable, and can be de-allocated.
+This set of objects includes all globals, as well as the objects which are reachable from the current stack frame.
+Any objects that are traversed during this operation are considered reachable and therefore cannot be deallocated.
+Conversely, any objects not traversed are no longer reachable, and can be deallocated.
 The other primary strategy for automated memory management, is reference counting garbage collection.
@@
 References are tracked by incrementing and decrementing an object's reference count when an assignment to a pointer occurs, or when a pointer variable goes out of scope.
-When the reference count reaches zero, it is safe to de-allocate the object.
+When the reference count reaches zero, it is safe to deallocate the object.
 
@@
 However, these pause times can still be an issue, especially in real time systems, where a hard guarantees on latency are needed.
-Reference counting has been discussed as a solution to this problem, but classic implementations of reference counting can have serious performance concerns when one de-allocation results in cascading de-allocations\cite{boehm}.
+Reference counting has been discussed as a solution to this problem since it incrementally deallocates objects at the program point where all references are destroyed, but classic implementations of reference counting can have serious performance concerns when one deallocation results in cascading deallocations\cite{boehm}.
 
@@
 In general a reference count garbage collector will have shorter pause times for the garbage collection phase, but will incur a higher performance penalty \cite{joisha}\cite{blackburn}.
-Reference-counting collectors have been implemented which defer the de-allocation, called lazy reference-counting, to some later time and some improvements on such strategies have been made \cite{boehm}.
+Reference-counting collectors have been implemented which defer the deallocation, called lazy reference-counting, to some later time and some improvements on such strategies have been made \cite{boehm}.
 Without extra work a strictly reference-counting collector will leak memory when references contain cycles, and multiple algorithms have been presented as solutions to this problem.
@@
 Prior to our modifications, the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs, with minimal memory constraints.
-After our modifications, the compiler properly maintains reference counts for all objects created, and automatically de-allocates objects when there are no longer any references to them.
+After our modifications, the compiler properly maintains reference counts for all objects created, and automatically deallocates objects when there are no longer any references to them.
 
@@
 \item Objects must include a counter that tracks the number of references to that object.
-\item Functions to increment and decrement an objects reference count must be created.
+\item Functions to increment and decrement an object's reference count must be created.
 \item Assignment of an object to a pointer or \textit{reference} should result in correct updates to an object's reference count.
@@
 However, the decrement function potentially has more work to do than just de-referencing the pointer and decrementing the reference count.
-According to the steps for reference counting mentioned previously, we need to de-allocate an object whenever its reference count reaches zero.
+According to the steps for reference counting mentioned previously, we need to deallocate an object whenever its reference count reaches zero.
 To provide this behavior, the \texttt{dec\_ref\_ctr} function has to check if the reference count has reached zero, and if so, delegate to a corresponding \texttt{free\_\textit{type}} method to free an object of the given type.
 
-In the \texttt{free\_\textit{type}} functions, it is necessary to add further calls to decrement the reference counter of any objects referenced by the object being de-allocated.
+In the \texttt{free\_\textit{type}} functions, it is necessary to add further calls to decrement the reference counter of any objects referenced by the object being deallocated.
 For the case of a list, we have the \texttt{free\_list} function, which iterates over each element in the list and decrements its reference count.
 For a dictionary, we have the \texttt{free\_dict} function, which decrements the reference counter on keys as well as values contained in the dictionary, since both can be objects.
-Note that for any case of an object that references other objects, this can can result in de-allocation of the referenced object, if the only reference to it is held by the parent object.
-In the worst case, this could lead to a number of de-allocations equivalent to the number of nodes in the object graph created by the parent object.
+Note that for any case of an object that references other objects, this can can result in deallocation of the referenced object, if the only reference to it is held by the parent object.
+In the worst case, this could lead to a number of deallocations equivalent to the number of nodes in the object graph created by the parent object.
 
@@
 For this reason, significant changes were made to the runtime to ensure that any type that references another type, has a generic pointer to a \texttt{big\_pyobj} structure, so that the reference counts can be correctly maintained.
-To ensure that we did not make any mistakes in this process, we create a runtime test suite that uses the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are de-allocated when their reference count reaches zero.
+To ensure that we did not make any mistakes in this process, we create a runtime test suite that uses the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are deallocated when their reference count reaches zero.
 This runtime test suite is important to unit test the runtime library, before attempting to use it from our compiler.
@@
 \item Ability to track allocations by type/size
-\item Ability to track the time of allocation/de-allocation
+\item Ability to track the time of allocation/deallocation
 \end{enumerate}
@@
 Every allocation of an object in our runtime is modified to use the \texttt{pymem\_new} and \texttt{pymem\_free} instead of \texttt{malloc} and \texttt{free}.
-As a result of this, we can determine exactly how much memory has been allocated and what objects have still not been de-allocated, by calling the \texttt{pymem\_print\_statistics} function at any given program point.
-In practice, this is only called at the end of each program to ensure that all memory has been de-allocated.
-Since we can track the time of allocation and de-allocation, this module can be beneficial for investigating object lifetime as well.
+As a result of this, we can determine exactly how much memory has been allocated and what objects have still not been deallocated, by calling the \texttt{pymem\_print\_statistics} function at any given program point.
+In practice, this is only called at the end of each program to ensure that all memory has been deallocated.
+Since we can track the time of allocation and deallocation, this module can be beneficial for investigating object lifetime as well.
 
@@
 The caller is then returned the portion of memory immediately after this pointer.
-Subsequent calls to \texttt{pymem\_free} then have access to the linked list node via the pointer, and can update the node to mark the memory as de-allocated, as well as the de-allocation time.
+Subsequent calls to \texttt{pymem\_free} then have access to the linked list node via the pointer, and can update the node to mark the memory as deallocated, as well as the deallocation time.
 The memory can then be freed independently of the linked list containing the statistics.
@@
 Each stage operates on an abstract syntax tree (AST).
-The relevant new stages are the pre-explicate flatten stage, and the reference counting AST transformation stage.
+The relevant new stages are the intermediate flatten stage, and the reference counting AST transformation stage.
 
@@
 
-\subsection{Challenges}
-We encountered a number of interesting challenges related to our design choices in modifying the compiler and summarize them below.
-
 \subsubsection{Expressions}
@@
 This leads to problems where memory is leaked, since no assignments ever occur to temporary objects.
-Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between to lists to the variable \texttt{a}.
-The assignment to the variable \texttt{a} results in a corresponding increment reference count on the result of the list addition.
+Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between to lists to the variable \texttt{x}.
+The assignment to the variable \texttt{x} results in a corresponding increment reference count on the result of the list addition.
 However, the temporary lists, \texttt{[1]} and \texttt{[2]} are leaked if we don't take any additional steps.
 To overcome this problem, we introduce a intermediate flattening stage, prior to our reference counting AST transformation.
-This results in the corresponding code given in Listing~\ref{lst:flattened}.
+This results in the corresponding code given in Listing~\ref{lst:flattened} after flattening.
+These temporary objects are then deallocated when their corresponding references go out of scope.
 
@@
 language=Python,
-frame=topline,
+frame=none,
 caption={A simple input program that leaks memory if not flattened.},
@@
 ]
-a = [1] + [2]
-a = 0
-print a
+x = [1] + [2]
+x = 0
+print x
 \end{lstlisting}
@@
 language=Python,
-frame=topline,
+frame=none,
 caption={The flattened version of listing \ref{lst:simpleleak}.},
@@
 tmp2 = tmp0 + tmp1
-a = tmp2
-a = 0
-print a
+x = tmp2
+x = 0
+print x
 \end{lstlisting}
@@
 \subsubsection{Temporary variables}
-As a result of introducing the intermediate flattening stage, we introduce additional references to an object which hang around until the end of the given program scope.
-This is undesirable because it changes the de-allocation behavior of the program from the expected behavior.
-For original input program in figure~\ref{lst:simpleleak}, the expectation is that when the variable \texttt{a} is assigned the value zero, the reference count for the list object should be decremented to zero and the object de-allocated.
+As a result of introducing the intermediate flattening stage, we introduce additional references to an object which hang around until the end of the given scope.
+This is undesirable because it changes the deallocation behavior of the program from the expected behavior.
+For original input program in figure~\ref{lst:simpleleak}, the expectation is that when the variable \texttt{x} is assigned the value zero, the reference count for the list object should be decremented to zero and the object deallocated.
 However, in the flattened program, this is no longer the case since the \texttt{tmp2} variable continues to reference the object.
-The de-allocation is therefore delayed until the end of the current scope.
+The deallocation is therefore delayed until the end of the current scope.
 To handle this situation, we keep track of all temporary variables introduced as a result of flattening an expression.
@@
 language=Python,
-frame=topline,
-caption={The flattened version of listing \ref{lst:simpleleak} with immediate de-allocation of temporary variables.},
+frame=none,
+caption={The flattened version of listing \ref{lst:simpleleak} with immediate deallocation of temporary variables.},
 captionpos=b,
@@
 tmp2 = tmp0 + tmp1
-a = tmp2
+x = tmp2
 tmp2 = 0
@@
 tmp0 = 0
-a = 0
-print a
+x = 0
+print x
 \end{lstlisting}
@@
 
-\subsection{Execution}
+We evaluated our reference counting implementation for correctness by using the \texttt{valgrind} utility against a simple application that uses a large number of language features.
+The program is given below.
 
-\subsection{Memory}
-
-\subsubsection{Valgrind results}
-For the trivial program
 \begin{verbatim}
@@
 list = []
-
 while i !=  10000:
@@
     i = i + 1
+\end{verbatim}
 
+Without any garbage collection, valgrind reports
+
+\footnotesize
+\begin{verbatim}
+LEAK SUMMARY:
+   definitely lost: 560,028 bytes in 20,001 blocks
+   indirectly lost: 199,004,992 bytes in 20,191 blocks
+     possibly lost: 1,375,008 bytes in 19,810 blocks
 \end{verbatim}
-Without any garbage collection, valgrind reports 
+\normalsize
+
+With our reference counting for the same program
+
+\footnotesize
 \begin{verbatim}
-==23467== LEAK SUMMARY:
-==23467==    definitely lost: 560,028 bytes in 20,001 blocks
-==23467==    indirectly lost: 199,004,992 bytes in 20,191 blocks
-==23467==      possibly lost: 1,375,008 bytes in 19,810 blocks
+LEAK SUMMARY:
+   definitely lost: 0 bytes in 0 blocks
+   indirectly lost: 0 bytes in 0 blocks
+     possibly lost: 0 bytes in 0 blocks
 \end{verbatim}
-With our reference counting for the same program
-\begin{verbatim}
-==23831== LEAK SUMMARY:
-==23831==    definitely lost: 0 bytes in 0 blocks
-==23831==    indirectly lost: 0 bytes in 0 blocks
-==23831==      possibly lost: 0 bytes in 0 blocks
-\end{verbatim}
+\normalsize
+
+\subsection{Reference counting overhead}
+
+To measure the overhead associated with reference counting, we run
+
+\subsection{De-allocation latency}
+
+The latency introduced by the \textit{dec\_ref\_ctr} is linear with the number of nodes in the object graph being deallocated.
+To illustrate this, the program given in listing~\ref{lst:tree} is run with different inputs, where the input is the height of a 4-ary tree.
+The number of nodes in such a tree is given by $\frac{{k}^{\left(h+1\right)}-1}{k-1}$ where $k=4$, and $h$ is the height of tree.
+The results are graphed in figure~\ref{fig:treelatency}.
+In typical programs, we expect the number of deallocations triggered by any given call to \textit{dec\_ref\_ctr} to be relatively small, but further evaluation of this for real world programs would be needed to prove this hypothesis.
+
+\begin{figure}[h!]
+\begin{lstlisting}[
+language=Python,
+frame=none,
+caption={An example input program that creates a tree of objects contained in lists.},
+captionpos=b,
+label=lst:tree
+]
+def fill(list, i, height):
+    if i != height:
+        j=0
+        while j != 4:
+            list[j]=[0,1,2,3]
+            fill(list[j], i+1, height)
+            j = j + 1
+    else:
+        0
+head=[0,1,2,3]
+fill(head, 0, input())
+\end{lstlisting}
+\end{figure}
+
+\begin{figure}
+\normalsize % use normal font size in the figure
+\input{treelatency.tex}
+\caption{Latency of \textit{dec\_ref\_ctr} for various size object graphs}
+\label{fig:treelatency}
+\end{figure}
 

### 2011-12-14 paper/paper.tex
Updates.

@@
 
-\conferenceinfo{WXYZ '05}{date, City.} 
+\conferenceinfo{CSCI5525 '11}{2011-Dec-15, Boulder, CO.} 
 \copyrightyear{2011} 
@@
 
-\title{Python to Assembly Compilation}
-\subtitle{With Reference Counting Garbage Collection}
+\title{An Implementation of Reference Counting for a Python to x86 Compiler}
+%\subtitle{With Reference Counting Garbage Collection}
 
@@
 frame=none,
+xleftmargin=0.2in,
+numbers=left,
 caption={A simple input program that leaks memory if not flattened.},
@@
 frame=none,
+xleftmargin=0.2in,
+numbers=left,
 caption={The flattened version of listing \ref{lst:simpleleak}.},
@@
 frame=none,
+xleftmargin=0.2in,
+numbers=left,
 caption={The flattened version of listing \ref{lst:simpleleak} with immediate deallocation of temporary variables.},
@@
 
-\subsubsection{Variable Assignment}
-We wanted a simple way to determine when to decrement an objects reference counter.
-Our first implementation was to add a call to decrement the counter before every assignment.
-However, because a variable can be created at any point in program execution we couldn't just blindly decrement the counter of an object since the variable will be uninitialized, and could be pointing at memory that makes it look like an object or accidentally references an object.
-\par
-The solution we took makes a slight modification to program behaviour, but we feel it is an acceptable trade-off.
-In every scope, the compiler adds an immediate assignment as the first instructions.
-This assignment is to a primitive, so the cost is minimal and then a decrement on the reference counter amounts to a function call with a no-op.
-This function call overhead could potentially be avoided by adding code to check if the variable points to an object or a primitive and is a future enhancement for performance.
-A program which used uninitialized (or declared) variables for program flow control would then have its runtime semantics changed.
-We felt this was a sufficient trade off since such use of an exception for program flow control is frowned upon in practice and the benefit to our design was an obvious positive.
+\subsubsection{Decrementing reference counts}
+As discussed previously, an assignment to a variable should result in a decrement of the reference count for the object referred to by the variable before the assignment.
+However, this can't be done when the variable doesn't refer to anything, which is true prior to the initial assignment.
+In fact, trying to even access such a variable before it is assigned to can result in undefined results, since there is no guarantee on the contents of the variable (it could be garbage or a value from a previous stack frame).
+To prevent a call to \texttt{dec\_ref\_ctr} on a variable before the initial assignment, we initially tried to statically determine when the first assignment occurred.
+This approach failed when we realized that for some programs, it is not possible to determine where the initial assignment occurs at compile-time.
+Listing~\ref{lst:initialassign} shows an example program where the initial assignment to the variable \texttt{x} cannot be determined statically.
+If the input provided by the user evaluates to \texttt{True}, then the initial assignment to $x$ occurs at line 2.
+However, if the input evaluates to \texttt{False}, then the initial assignment to \texttt{x} occurs at line 3.
+If we transform the input program so that a call is made to decrement the reference count for \texttt{x} immediately prior to the assignment on line 3, then there is the possibility that this will cause the program to crash if the user provides an input that evaluates to \texttt{False}.
 
-\subsubsection{Addition of Lists}
-In the current runtime an entirely new list is created when two lists are added together.
-At the assignment level our compiler can determine (due to the flattened structure) that the left-hand-side and the right-hand-side of the addition need their reference counters manipulated, however the assignment from an addition to a list creates an entirely new list which should properly increment the reference counter to all objects in both lists.
+To solve this issue, we make a simple change.
+We ensure that the value of \texttt{x} is always initialized, by assigning it the value zero at the program scope where it is defined\footnote{Python defines this as the block where a variable is initially assigned to.}.
+In our compiler, this is equivalent to a primitive of type \texttt{int}, but this can be any sentinel value that results in a no-op when a call is made to \texttt{dec\_ref\_ctr}.
+By ensuring that all local variables are assigned a sentinel value at the beginning of the scope where they are defined, we can safely transform our input program to call \texttt{dec\_ref\_ctr} immediately preceding an assignment, and \texttt{inc\_ref\_ctr} immediately succeeding an assignment.
+
+This solution is somewhat controversial, since it can alter program behavior when the program tries to read from a variable that has not yet been assigned to.
+The standard Python interpreter throws a \textit{NameError} in this case, but our compiler will allow such a program to be compiled.
+Reading the variable will then result in the sentinel value (zero in our case) being read, instead of the program crashing or having undefined behavior.
+A program which used uninitialized (or declared) variables would then have its runtime behavior changed.
+Another option to address this issue, is to make it a compile-time error to try and access variable from a conditional block that may have not been previously initialized.
+This is similar to a restriction in the Java language.
+
+\begin{figure}[h!]
+\begin{lstlisting}[
+language=Python,
+frame=none,
+xleftmargin=0.2in,
+numbers=left,
+caption={An example program where the initial assignment to \texttt{x} cannot be determined statically.},
+captionpos=b,
+label=lst:initialassign
+]
+if input():
+    x = [1]
+x = [2]
+\end{lstlisting}
+\end{figure}
 
 \subsubsection{Functions}
-
 The compiler creates temporary variables in a number of situations (namely function calls and as a result of flattened expressions).
@@
 
-\subsection{Assembly Output}
-We modified the x86 assembly to include calls to set up our memory tracking system and print statistics at the end of the program.
-Running analysis tools such as valgrind on our programs before and after shows the obvious gain of going from every memory allocation leaks to zero leaks.
+\subsubsection{Addition of Lists}
+In the current runtime an entirely new list is created when two lists are added together.
+At the assignment level our compiler can determine (due to the flattened structure) that the left-hand-side and the right-hand-side of the addition need their reference counters manipulated, however the assignment from an addition to a list creates an entirely new list which should properly increment the reference counter to all objects in both lists.
+
+\subsubsection{Debugging Integration}
+To facilitate debugging in our compiler, we modify the x86 assembly generator to include calls to set up our \texttt{pymem} module and print statistics at the end of the program.
 
@@
 \label{sec:results}
-
 We evaluated our reference counting implementation for correctness by using the \texttt{valgrind} utility against a simple application that uses a large number of language features.
@@
 frame=none,
+xleftmargin=0.2in,
+numbers=left,
 caption={An example input program that creates a tree of objects contained in lists.},

### 2011-12-14 compiler/p0generator.py
- Fix code to measure dec_ref_latency by adding a runtime_init and runtime_shutdown   method.  The maximum latency will be printed on the call to runtime_shutdown. - Fix a minor issue with the quickcompile.sh script.

@@ -36,8 +36,10 @@ main:
 \tmovl %%esp, %%ebp
 \tsubl $%s,%%esp # make stack space for variables
 \tcall pymem_init
+\tcall runtime_init
 
 %s
+\tcall runtime_shutdown
 \tcall pymem_print_stats
 \tcall pymem_shutdown
 \tmovl $0, %%eax # put return value in eax

### 2011-12-14 compiler/quickcompile.sh

@@ -6,9 +6,12 @@ if [ "$#" -ne "1" ] ; then
   exit 1
 fi
 
-PYFILE=$1
+PYFILE=`basename $1`
 BASENAME=`echo "$PYFILE" | cut -f1 -d.`
 SFILE="${BASENAME}.s"
 
-./compile.py $PYFILE
+echo ./compile.py "$PYFILE"
+./compile.py "$PYFILE"
+
+echo gcc -o "$BASENAME" "$SFILE" -lm *.c -pg -m32
 gcc -o "$BASENAME" "$SFILE" -lm *.c -pg -m32

### 2011-12-14 compiler/runtime.c

@@ -12,6 +12,8 @@ int min(int x, int y) {
     return y < x ? y : x;
 }
 
+static struct timeval decref_latency;
+
 /* Some forward declarations */
 static int equal_pyobj(pyobj a, pyobj b);
 static void print_float(double in);
@@ -1397,14 +1399,34 @@ void dec_ref_ctr(pyobj v) {
     struct timeval start, end, result;
     int ret;
 
-    //ret = gettimeofday(&start, NULL);
-    //assert (ret > -1);
+    ret = gettimeofday(&start, NULL);
+    assert (ret > -1);
     dec_ref_ctr_rec(v);
-    //ret = gettimeofday(&end, NULL);
-    //assert (ret > -1);
-    //timeval_subtract(&result, &end, &start);
-    //fprintf(stderr, "dec_ref_ctr took %ld.%ld seconds\n", result.tv_sec, result.tv_usec);
+    ret = gettimeofday(&end, NULL);
+    assert (ret > -1);
+    timeval_subtract(&result, &end, &start);
+
+    if (result.tv_sec > decref_latency.tv_sec || 
+        (result.tv_sec == decref_latency.tv_sec &&
+         result.tv_usec > decref_latency.tv_usec)
+       )
+    {
+        decref_latency.tv_sec = result.tv_sec;
+        decref_latency.tv_usec = result.tv_usec;
+    }
 }
+
+void runtime_init()
+{
+    decref_latency.tv_sec = 0;
+    decref_latency.tv_usec = 0;
+}
+
+void runtime_shutdown()
+{
+    fprintf(stderr, "dec_ref_ctr took %ld.%06ld seconds\n", decref_latency.tv_sec, decref_latency.tv_usec);
+}
+
 
 /* Subtract the `struct timeval' values X and Y,
    storing the result in RESULT.
@@ -1427,7 +1449,7 @@ int timeval_subtract (struct timeval *result, struct timeval *x, struct timeval 
        tv_usec is certainly positive. */
     result->tv_sec = x->tv_sec - y->tv_sec;
     result->tv_usec = x->tv_usec - y->tv_usec;
-     
+
     /* Return 1 if result is negative. */
     return x->tv_sec < y->tv_sec;
 }

### 2011-12-14 compiler/runtime.h

@@ -153,5 +153,7 @@ pyobj error_pyobj(char* string);
 
 void inc_ref_ctr(pyobj v);
 void dec_ref_ctr(pyobj v);
+void runtime_init();
+void runtime_shutdown();
 
 #endif /* RUNTIME_H */

### 2011-12-14 compiler/gcrefcount.py
- Add autorelease functionality for handling reference counting for   function return values.

@@ -66,7 +66,7 @@ class GCRefCount:
         self.varalloc = varalloc
         self.varset = set()
         self.log = logging.getLogger('compiler.gcrefcounter')
-        self.lambda_local_assigns = []
+        self.lambda_local_vars = []
     
     def transform(self, node, *args, **kwargs):
         self.log.info('Starting gcrefcounter')
@@ -117,8 +117,7 @@ class GCRefCount:
             self.varset.add(assnode.name)
             stmtlist.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(assnode.name)])))
             stmtlist.append(Assign([AssName(assnode.name,'OP_ASSIGN')], self.visit(node.expr)))
-            if not isinstance(node.expr,CallFuncIndirect):
-                stmtlist.append(Discard(CallFunc(Name('inc_ref_ctr'),[Name(assnode.name)])))
+            stmtlist.append(Discard(CallFunc(Name('inc_ref_ctr'),[Name(assnode.name)])))
             return stmtlist
         else:
             raise Exception('Unknown assignment node: %s' % assnode)
@@ -171,13 +170,17 @@ class GCRefCount:
         return node
     
     def visit_Return(self, node, *args, **kwargs):
-        localAssigns = self.lambda_local_assigns[-1]
+        localAssigns, argset = self.lambda_local_vars[-1]
         self.log.debug('visit_Return: localAssigns=%s' % localAssigns)
         self.log.debug('visit_Return: node=%s' % node)
         decrefstmts=[]
-        for localvar in localAssigns:
-            if not isinstance(node.value, Name) or node.value.name not in localAssigns:
+        # call dec_ref on all local variables, unless the this value is being returned.
+        for localvar in localAssigns | argset:
+            if not isinstance(node.value, Name) or node.value.name != localvar:
                 decrefstmts.append(Discard(CallFunc(Name('dec_ref_ctr'),[Name(localvar)])))
+        # Call set_autorelease on the variable being returned from the function
+        if isinstance(node.value,Name):
+            decrefstmts.append(Discard(CallFunc(Name('autorelease'),[node.value])))
         return decrefstmts + [node]
 
     def visit_CallFuncIndirect(self, node, *args, **kwargs):
@@ -211,6 +214,7 @@ class GCRefCount:
         # GCFlattener; recurse on that stmt object here to add ref counting.
         if not isinstance(node.code, Stmt):
             raise Exception('Expected Stmt object for Lambda.code')
+        increfstmts = []
[…]

### 2011-12-14 compiler/runtime.c

@@ -243,6 +243,7 @@ static big_pyobj* list_to_big(list l) {
     v->tag = LIST;
     v->u.l = l;
     v->ref_ctr = 0;
+    v->deferred_cnt = 0;
     return v;
 }
 
@@ -550,6 +551,7 @@ big_pyobj* create_dict()
     v->tag = DICT;
     v->u.d = create_hashtable(4, hash_any, equal_any, DICT);
     v->ref_ctr = 0;
+    v->deferred_cnt = 0;
     return v;
 }
 
@@ -837,6 +839,7 @@ static big_pyobj* closure_to_big(function f) {
     v->tag = FUN;
     v->u.f = f;
     v->ref_ctr = 0;
+    v->deferred_cnt = 0;
     return v;
 }
 
@@ -970,6 +973,7 @@ big_pyobj* create_class(pyobj bases)
     ret = (big_pyobj*)pymem_new(CLASS, sizeof(big_pyobj));
     ret->tag = CLASS;
     ret->ref_ctr = 0;
+    ret->deferred_cnt = 0;
     ret->u.cl.attrs = create_hashtable(2, attrname_hash, attrname_equal, CLASS);
     ret->u.cl.parents = bases;
     // remember to increment the reference count on the list pyobj of base classes!
@@ -982,6 +986,7 @@ big_pyobj* create_object(pyobj cl) {
     big_pyobj* ret = (big_pyobj*)pymem_new(OBJECT, sizeof(big_pyobj));
     ret->tag = OBJECT;
     ret->ref_ctr = 0;
+    ret->deferred_cnt = 0;
     big_pyobj* clp = project_big(cl);
     if (clp->tag == CLASS) {
         ret->u.obj.clazz = cl;
@@ -991,7 +996,6 @@ big_pyobj* create_object(pyobj cl) {
         exit(-1);
     }
     ret->u.obj.attrs = create_hashtable(2, attrname_hash, attrname_equal, OBJECT);
-    inc_ref_ctr(inject_big(ret));
     return ret;
 }
 
@@ -1028,6 +1032,7 @@ static big_pyobj* create_bound_method(pyobj receiver, pyobj f) {
     big_pyobj* ret = (big_pyobj*)pymem_new(BMETHOD, sizeof(big_pyobj));
     ret->tag = BMETHOD;
     ret->ref_ctr = 0;
+    ret->deferred_cnt = 0;
     ret->u.bm.fun = f;
     inc_ref_ctr(f);
[…]

### 2011-12-14 compiler/runtime.h

@@ -97,6 +97,7 @@ struct pyobj_struct {
         bound_method bm;
     } u;
     int ref_ctr;
+    int deferred_cnt;
 };
 typedef struct pyobj_struct big_pyobj;
 
@@ -153,6 +154,7 @@ pyobj error_pyobj(char* string);
 
 void inc_ref_ctr(pyobj v);
 void dec_ref_ctr(pyobj v);
+void autorelease(pyobj v);
 void runtime_init();
 void runtime_shutdown();
 

### 2011-12-14 compiler/test/funcdoubleret.py

@@ -1,0 +1,6 @@
+def f(x):
+    return x
+
+a = f(f([1]))
+print a
+print a

### 2011-12-14 compiler/test/tree.py

@@ -1,0 +1,15 @@
+# vim: set ts=4 sw=4 expandtab:
+def fill(list, i, height):
+    if i != height:
+        j=0
+        while j != 4:
+            list[j]=[0,1,2,3]
+            fill(list[j], i+1, height)
+            j = j + 1
+    else:
+        0
+
+head=[0,1,2,3]
+fill(head, 0, input())
+print 777
+input()

### 2011-12-14 compiler/runtime.c
- Fix print_dict so that is properly reference counts the   printint_dict variable and any temporary variables. - Do not call dec_ref_ctr_rec from dec_ref_ctr, if not a big.

@@ -168,12 +168,12 @@ static void print_bmethod(pyobj obj) {
     big_pyobj* b = project_big(obj);
     printf("bound method receiver ");
     print_pyobj(b->u.bm.receiver);
-    printf("    and fun at address %X\n", b->u.bm.fun);
+    printf("    and fun at address %p\n", (void *)b->u.bm.fun);
 }
 
 static void print_fun(pyobj obj) {
     big_pyobj* b = project_big(obj);
-    printf("function at address %X\n", b->u.f.function_ptr);
+    printf("function at address %p\n", b->u.f.function_ptr);
 }
 
 static void print_pyobj(pyobj x) {
@@ -290,7 +290,7 @@ static int list_equal(list x, list y)
 */
 
 static char inside;
-static list printing_list;
+static big_pyobj *printing_list = NULL;
 
 static void print_dict(pyobj dict)
 {
@@ -299,12 +299,12 @@ static void print_dict(pyobj dict)
     if(!inside) {
         inside = 1;
         inside_reset = 1;
-        printing_list.len = 0;
-        printing_list.data = 0;
+        printing_list = create_list(inject_int(0));
+        inc_ref_ctr(inject_big(printing_list));
     }
     d = project_big(dict);
 
-    if(is_in_list(printing_list, dict)) {
+    if(is_in_list(printing_list->u.l, dict)) {
         printf("{...}");
         return;
     }
@@ -319,18 +319,23 @@ static void print_dict(pyobj dict)
             pyobj v = *(pyobj *)hashtable_iterator_value(itr);
             print_pyobj(k);
             printf(": ");
-            if (is_in_list(printing_list, v)
+            if (is_in_list(printing_list->u.l, v)
                     || equal_pyobj(v,dict)) {
                 printf("{...}");
             }
             else {
                 /* tally this dictionary in our list of printing dicts */
-                list a;
-                a.len = 1;
-                a.data = (pyobj*)pymem_new(LIST, sizeof(pyobj) * a.len);
-                a.data[0] = dict;
+                big_pyobj *list;
+                big_pyobj *new_printing_list;
+                list = create_list(inject_int(1));
+                inc_ref_ctr(inject_big(list));
+                set_subscript(inject_big(list), inject_int(0), dict);
[…]

### 2011-12-14 compiler/test/dict2.py

@@ -1,0 +1,2 @@
+a = {1:[{2:3}]}
+print a

### 2011-12-14 compiler/quickcompile.sh
- Fix runtime_tests, which were broken when we changed the prototype of   inc_ref_ctr and dec_ref_ctr. - Fix all tests to print out something at the end, otherwise we might be silently   failing tests. - Add allfeatures.py and ctor.py test cases.  Constructors are broken and ctor.py   will fail.  But allfeatures.py should succeed.

@@ -13,5 +13,5 @@ SFILE="${BASENAME}.s"
 echo ./compile.py "$PYFILE"
 ./compile.py "$PYFILE"
 
-echo gcc -o "$BASENAME" "$SFILE" -lm *.c -pg -m32
-gcc -o "$BASENAME" "$SFILE" -lm *.c -pg -m32
+echo gcc -g -o "$BASENAME" "$SFILE" *.c -lm -m32
+gcc -g -o "$BASENAME" "$SFILE" *.c -lm -m32

### 2011-12-14 compiler/runtime_test/closure.c

@@ -10,6 +10,9 @@
  * infrastructure.
  */
 
+void incref (big_pyobj *obj) { inc_ref_ctr(inject_big(obj)); }
+void decref (big_pyobj *obj) { dec_ref_ctr(inject_big(obj)); }
+
 int dummy_function (int x)
 {
     return x;
@@ -28,16 +31,16 @@ int main (int argc, char *argv[])
      */
     pymem_init();
     freevars = create_list(inject_int(0));
-    inc_ref_ctr(freevars);
+    incref(freevars);
 
     f = create_closure(dummy_function, inject_big(freevars));
-    inc_ref_ctr(f);
+    incref(f);
     assert (freevars->ref_ctr == 2);
 
-    dec_ref_ctr(f);
+    decref(f);
     assert (freevars->ref_ctr == 1);
 
-    dec_ref_ctr(freevars);
+    decref(freevars);
 
     pymem_print_stats();
     pymem_shutdown();

### 2011-12-14 compiler/runtime_test/dict.c

@@ -9,6 +9,9 @@
  * infrastructure.
  */
 
+void incref (big_pyobj *obj) { inc_ref_ctr(inject_big(obj)); }
+void decref (big_pyobj *obj) { dec_ref_ctr(inject_big(obj)); }
+
 int main (int argc, char *argv[])
 {
     big_pyobj *dict, *dict2;
@@ -19,8 +22,8 @@ int main (int argc, char *argv[])
      */
     pymem_init();
     dict = create_dict();
-    inc_ref_ctr(dict);
-    dec_ref_ctr(dict);
+    incref(dict);
+    decref(dict);
     pymem_print_stats();
     pymem_shutdown();
 
@@ -29,9 +32,9 @@ int main (int argc, char *argv[])
      */
     pymem_init();
     dict = create_dict();
-    inc_ref_ctr(dict);
+    incref(dict);
     set_subscript(inject_big(dict), inject_int(1), inject_int(2));
-    dec_ref_ctr(dict);
+    decref(dict);
     pymem_print_stats();
     pymem_shutdown();
 
@@ -39,18 +42,18 @@ int main (int argc, char *argv[])
      * test3: dictionary with one pyobj entry for the value
      */
     pymem_init();
-    dict = create_dict(); inc_ref_ctr(dict);
-    dict2 = create_dict(); inc_ref_ctr(dict2);
+    dict = create_dict(); incref(dict);
+    dict2 = create_dict(); incref(dict2);
 
     set_subscript(inject_big(dict), inject_int(1), inject_big(dict2));
 
     // verify the ref count for dict2 is now 2
     assert (dict2->ref_ctr == 2);
-    dec_ref_ctr(dict);
+    decref(dict);
 
     // at this point, dict should no longer exist, but dict2 should still exist
     assert (dict2->ref_ctr == 1);
-    dec_ref_ctr(dict2);
+    decref(dict2);
 
     pymem_print_stats();
     pymem_shutdown();

### 2011-12-14 compiler/runtime_test/list.c

@@ -10,6 +10,10 @@
  * infrastructure.
  */
 
+void incref (big_pyobj *obj) { inc_ref_ctr(inject_big(obj)); }
+void decref (big_pyobj *obj) { dec_ref_ctr(inject_big(obj)); }
+
+
 int main (int argc, char *argv[])
 {
     big_pyobj *list, *list2;
@@ -23,8 +27,8 @@ int main (int argc, char *argv[])
      */
     pymem_init();
     list = create_list(inject_int(0));
-    inc_ref_ctr(list);
-    dec_ref_ctr(list);
+    incref(list);
+    decref(list);
     pymem_print_stats();
     pymem_shutdown();
 
@@ -35,13 +39,13 @@ int main (int argc, char *argv[])
      */
     pymem_init();
     list = create_list(inject_int(5));
-    inc_ref_ctr(list);
+    incref(list);
     set_subscript(inject_big(list), inject_int(0), inject_int(99));
     set_subscript(inject_big(list), inject_int(1), inject_int(88));
     set_subscript(inject_big(list), inject_int(2), inject_int(77));
     set_subscript(inject_big(list), inject_int(3), inject_int(66));
     set_subscript(inject_big(list), inject_int(4), inject_int(55));
-    dec_ref_ctr(list);
+    decref(list);
     pymem_print_stats();
     pymem_shutdown();
 
@@ -56,18 +60,18 @@ int main (int argc, char *argv[])
     pymem_init();
 
     list = create_list(inject_int(1));
-    inc_ref_ctr(list);
+    incref(list);
     set_subscript(inject_big(list), inject_int(0), inject_int(99));
 
     list2 = create_list(inject_int(1));
-    inc_ref_ctr(list2);
+    incref(list2);
     set_subscript(inject_big(list2), inject_int(0), inject_big(list));
     assert (list->ref_ctr == 2);
 
-    dec_ref_ctr(list2);
+    decref(list2);
     assert (list->ref_ctr == 1);
 
-    dec_ref_ctr(list);
+    decref(list);
     pymem_print_stats();
     pymem_shutdown();
 }

### 2011-12-14 compiler/runtime_test/object.c

@@ -10,6 +10,9 @@
  * work correctly to release allocated memory. Relies on the pymem
  * infrastructure.
  */
+void incref (big_pyobj *obj) { inc_ref_ctr(inject_big(obj)); }
+void decref (big_pyobj *obj) { dec_ref_ctr(inject_big(obj)); }
+
 
 int main (int argc, char *argv[])
 {
@@ -26,24 +29,24 @@ int main (int argc, char *argv[])
      */
     pymem_init();
     A_parents = create_list(inject_int(0));
-    inc_ref_ctr(A_parents);
+    incref(A_parents);
 
     A = create_class(inject_big(A_parents));
-    inc_ref_ctr(A);
+    incref(A);
     assert (A_parents->ref_ctr == 2);
 
     a = create_object(inject_big(A));
-    inc_ref_ctr(a);
+    incref(a);
     assert (A->ref_ctr == 2);
 
-    dec_ref_ctr(A_parents);
+    decref(A_parents);
     assert (A_parents->ref_ctr == 1);
 
-    dec_ref_ctr(A);
+    decref(A);
     assert (A->ref_ctr == 1);
 
     pymem_print_stats();
-    dec_ref_ctr(a);   // this should deallocate everything
+    decref(a);   // this should deallocate everything
     pymem_print_stats();
     pymem_shutdown();
 
@@ -64,49 +67,49 @@ int main (int argc, char *argv[])
 
     pymem_init();
     A_parents = create_list(inject_int(0));
-    inc_ref_ctr(A_parents);
+    incref(A_parents);
 
     A = create_class(inject_big(A_parents));
-    inc_ref_ctr(A);
+    incref(A);
     assert (A_parents->ref_ctr == 2);
 
     B_parents = create_list(inject_int(1));
-    inc_ref_ctr(B_parents);
+    incref(B_parents);
     set_subscript(inject_big(B_parents), inject_int(0), inject_big(A));
     assert (A->ref_ctr == 2);
 
     B = create_class(inject_big(B_parents));
-    inc_ref_ctr(B);
+    incref(B);
     assert (B_parents->ref_ctr == 2);
 
     a = create_object(inject_big(A));
-    inc_ref_ctr(a);
+    incref(a);
     assert (A->ref_ctr == 3);
 
     b = create_object(inject_big(B));
-    inc_ref_ctr(b);
+    incref(b);
     assert (B->ref_ctr == 2);
 
[…]

### 2011-12-14 compiler/test/allfeatures.py

@@ -1,0 +1,33 @@
+a = {1:2}
+print a
+a = 0
+
+class B:
+    x = 1
+
+print B.x
+b = B()
+b.y = 2
+print b.y
+b = 0
+
+class C(B):
+    y = 2
+c = C()
+c.z = 3
+print c.x
+print c.y
+print c.z
+
+def f(x,y):
+    g = lambda x: x
+    i = 0
+    lst = []
+    while i !=  10000:
+        lst = lst + [[i]]
+        i = i + 1
+    return g(lst)
+
+print f(1,2)
+
+

### 2011-12-14 compiler/test/func.py

@@ -1,3 +1,4 @@
 def f():
     False
 f()
+print 1

### 2011-12-14 compiler/test/func2.py

@@ -1,3 +1,4 @@
 def f():
     []
 f()
+print 1

### 2011-12-14 compiler/test/lambda.py

@@ -1,2 +1,2 @@
 f = lambda: False
-f()
+print f()

### 2011-12-14 compiler/test/nested.py

@@ -1,3 +1,3 @@
 def f():
     return lambda: 1 + 2
-f()
+print f()()

### 2011-12-14 compiler/test/ref1.py

@@ -1,4 +1,6 @@
 tmp1 = [1] + [2]
 input()
+print tmp1
 tmp1 = 0
 input()
+print tmp1

### 2011-12-14 compiler/test/ref2.py

@@ -5,3 +5,4 @@ input()
 tmp1 = 0
 input()
 f()
+print tmp1

### 2011-12-14 compiler/test/ref3.py

@@ -1,4 +1,6 @@
 test = [] + []
 input()
+print test
 test = 0
 input()
+print test

### 2011-12-14 compiler/test/simplelist.py

@@ -3,3 +3,4 @@ a[0]=[0,1,2,3]
 a[1]=[0,1,2,3]
 a[2]=[0,1,2,3]
 a[3]=[0,1,2,3]
+print a

### 2011-12-15 paper/paper.tex
Almost done.

@@
 \usepackage{listings}
+\usepackage{subfigure}
 
+% Define the style for listings
 \lstset{
@@
 numberstyle=\small,
-commentstyle=\textit
+commentstyle=\sffamily\textit,
+columns=fullflexible,
 stepnumber=1
 }
+
+% some macros for consistency
+\newcommand{\avariable}[1]{\ensuremath{#1}}
+\newcommand{\aliteral}[1]{\texttt{#1}}
+\newcommand{\afunction}[1]{\textit{#1}}
+\newcommand{\atype}[1]{\texttt{#1}}
+\newcommand{\afield}[1]{\texttt{#1}}
 
@@
 Section~\ref{sec:results} evaluates our compiler for performance and correctness.
-Section~\ref{sec:future} discusses potential future work.
 Finally, section~\ref{sec:conclusion} concludes.
@@
 Our runtime library is implemented in C so that it can be easily called by the assembly programs created by our compiler.
-The runtime library contains abstractions such as \texttt{create\_list}, \texttt{create\_dict} and \texttt{create\_object} that take care of allocating and initializing memory for a new object of the specified type.
+The runtime library contains abstractions such as \afunction{create\_list}, \afunction{create\_dict} and \afunction{create\_object} that take care of allocating and initializing memory for a new object of the specified type.
 These functions return a pointer to the memory allocated, and the pointer is \textit{tagged} to indicate that it represents a non-primitive type.
-In our reference compiler, any non-primitive object must be a pointer to a C structure called \texttt{big\_pyobj}.
-To support reference counts, this C structure is modified to include an int field called \texttt{ref\_ctr}.
+In our reference compiler, any non-primitive object must be a pointer to a C structure called \atype{big\_pyobj}.
+To support reference counts, this C structure is modified to include an int field called \afield{ref\_ctr}.
 Then all of the functions to create objects are modified to set the initial reference count to zero.
 
-To support incrementing and decrementing an object's reference count, we create corresponding functions in our runtime called \texttt{inc\_ref\_ctr} and \texttt{dec\_ref\_ctr}.
-The increment function is trivial in that it simply de-references the object pointer and increments the \texttt{ref\_ctr} field.
+To support incrementing and decrementing an object's reference count, we create corresponding functions in our runtime called \afunction{inc\_ref\_ctr} and \afunction{dec\_ref\_ctr}.
+The increment function is trivial in that it simply de-references the object pointer and increments the \afield{ref\_ctr} field.
 However, the decrement function potentially has more work to do than just de-referencing the pointer and decrementing the reference count.
 According to the steps for reference counting mentioned previously, we need to deallocate an object whenever its reference count reaches zero.
-To provide this behavior, the \texttt{dec\_ref\_ctr} function has to check if the reference count has reached zero, and if so, delegate to a corresponding \texttt{free\_\textit{type}} method to free an object of the given type.
+To provide this behavior, the \afunction{dec\_ref\_ctr} function has to check if the reference count has reached zero, and if so, delegate to a corresponding \afunction{free\_\textit{type}} method to free an object of the given type.
 
-In the \texttt{free\_\textit{type}} functions, it is necessary to add further calls to decrement the reference counter of any objects referenced by the object being deallocated.
-For the case of a list, we have the \texttt{free\_list} function, which iterates over each element in the list and decrements its reference count.
-For a dictionary, we have the \texttt{free\_dict} function, which decrements the reference counter on keys as well as values contained in the dictionary, since both can be objects.
+In the \afunction{free\_\textit{type}} functions, it is necessary to add further calls to decrement the reference counter of any objects referenced by the object being deallocated.
+For the case of a list, we have the \afunction{free\_list} function, which iterates over each element in the list and decrements its reference count.
+For a dictionary, we have the \afunction{free\_dict} function, which decrements the reference counter on keys as well as values contained in the dictionary, since both can be objects.
 Note that for any case of an object that references other objects, this can can result in deallocation of the referenced object, if the only reference to it is held by the parent object.
@@
 Since every object in our system may contain references to another object, we modify our runtime to correctly increment the reference count when such as reference is established.
-For example, an assignment to a list or dictionary subscript results in a call to the \texttt{set\_subscript} runtime function.
+For example, an assignment to a list or dictionary subscript results in a call to the \afunction{set\_subscript} runtime function.
 This function is modified to first decrement the reference count for any object that may have been previously referenced by the given subscript, and then increment the reference count for the new object that is assigned to the given subscript.
@@
 
-In some cases, the type definition for an object in our runtime did not aggregate a \texttt{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \texttt{class\_struct}).
-For this reason, significant changes were made to the runtime to ensure that any type that references another type, has a generic pointer to a \texttt{big\_pyobj} structure, so that the reference counts can be correctly maintained.
+In some cases, the type definition for an object in our runtime did not aggregate a \atype{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \atype{class\_struct}).
+For this reason, significant changes were made to the runtime to ensure that any type that references another type, has a generic pointer to a \atype{big\_pyobj} structure, so that the reference counts can be correctly maintained.
 To ensure that we did not make any mistakes in this process, we create a runtime test suite that uses the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are deallocated when their reference count reaches zero.
@@
 
-Finally, in order to verify the correctness of our implementation, we create a set of abstractions on top of the system allocation functions \texttt{malloc} and \texttt{free}.
+Finally, in order to verify the correctness of our implementation, we create a set of abstractions on top of the system allocation functions \afunction{malloc} and \afunction{free}.
 These abstractions are part of our runtime and provide us the following benefits.
@@
 The abstractions are implemented in a new module appropriately called \texttt{pymem}.
-Every allocation of an object in our runtime is modified to use the \texttt{pymem\_new} and \texttt{pymem\_free} instead of \texttt{malloc} and \texttt{free}.
-As a result of this, we can determine exactly how much memory has been allocated and what objects have still not been deallocated, by calling the \texttt{pymem\_print\_statistics} function at any given program point.
-In practice, this is only called at the end of each program to ensure that all memory has been deallocated.
+Every allocation of an object in our runtime is modified to use the \afunction{pymem\_new} and \afunction{pymem\_free} instead of \afunction{malloc} and \afunction{free}.
+As a result of this, we can determine exactly how much memory has been allocated and what objects have still not been deallocated, by calling the \afunction{pymem\_print\_statistics} function at any given program point.
+In practice, this is only called at the end of each program\footnote{This is achieved by modifying the x86 generation phase of our compiler to include calls to set up our \texttt{pymem} module and print statistics at the end of the program.
+} to ensure that all memory has been deallocated.
 Since we can track the time of allocation and deallocation, this module can be beneficial for investigating object lifetime as well.
 
-The \texttt{pymem\_new} routine works by allocating slightly more memory than is requested by the user.
-This extra memory is used to store a pointer to a linked list node that contains the details of the allocation, including the type of object being created, the requested size, the pointer returned by \texttt{malloc} and the time of allocation.
+The \afunction{pymem\_new} routine works by allocating slightly more memory than is requested by the user.
+This extra memory is used to store a pointer to a linked list node that contains the details of the allocation, including the type of object being created, the requested size, the pointer returned by \afunction{malloc} and the time of allocation.
 The caller is then returned the portion of memory immediately after this pointer.
-Subsequent calls to \texttt{pymem\_free} then have access to the linked list node via the pointer, and can update the node to mark the memory as deallocated, as well as the deallocation time.
+Subsequent calls to \afunction{pymem\_free} then have access to the linked list node via the pointer, and can update the node to mark the memory as deallocated, as well as the deallocation time.
 The memory can then be freed independently of the linked list containing the statistics.
@@
 Since we only want to decrement reference counts for \textit{tagged} types\footnote{Actually, we only want to modify reference counts for a subset of tagged types that only includes ``big'' types.
-However, to simplify our implementation, we transform our AST to always call our increment and decrement runtime functions, but these functions will ignore the primitive types of \texttt{int} and \texttt{boolean}.}, the only option is to put this transformation stage before type explication.
+However, to simplify our implementation, we transform our AST to always call our increment and decrement runtime functions, but these functions will ignore the primitive types of \atype{int} and \atype{boolean}.}, the only option is to put this transformation stage before type explication.
 In addition to modifying the AST prior before/after an assignment, the reference counting transformation is also responsible for determining which variables are local to a given scope and decrementing the counts for these references at the end of the given program scope.
@@
 This leads to problems where memory is leaked, since no assignments ever occur to temporary objects.
-Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between to lists to the variable \texttt{x}.
-The assignment to the variable \texttt{x} results in a corresponding increment reference count on the result of the list addition.
-However, the temporary lists, \texttt{[1]} and \texttt{[2]} are leaked if we don't take any additional steps.
+Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between to lists to the variable \avariable{x}.
+The assignment to the variable \avariable{x} results in a corresponding increment reference count on the result of the list addition.
+However, the temporary lists, \aliteral{[1]} and \aliteral{[2]} are leaked if we don't take any additional steps.
 To overcome this problem, we introduce a intermediate flattening stage, prior to our reference counting AST transformation.
@@
 
-\begin{figure}[h!]
+\begin{figure}[h]
 \begin{lstlisting}[
@@
 numbers=left,
-caption={A simple input program that leaks memory if not flattened.},
+caption={Program that leaks memory if not flattened.},
 captionpos=b,
@@
 This is undesirable because it changes the deallocation behavior of the program from the expected behavior.
-For original input program in figure~\ref{lst:simpleleak}, the expectation is that when the variable \texttt{x} is assigned the value zero, the reference count for the list object should be decremented to zero and the object deallocated.
-However, in the flattened program, this is no longer the case since the \texttt{tmp2} variable continues to reference the object.
+For original input program in listing~\ref{lst:simpleleak}, the expectation is that when the variable \avariable{x} is assigned the value zero, the reference count for the list object should be decremented to zero and the object deallocated.
+However, in the flattened program, this is no longer the case since the \avariable{tmp2} variable continues to reference the object.
 The deallocation is therefore delayed until the end of the current scope.
@@
 This will lead to an immediate decrement of the reference count after the reference counting AST transformation.
-Figure~\ref{lst:fixflattened} shows the result of the flattening transformation after this change is made.
+Listing~\ref{lst:fixflattened} shows the result of the flattening transformation after this change is made.
 
-\begin{figure}[h!]
+\begin{figure}[h]
 \begin{lstlisting}[
@@
 numbers=left,
-caption={The flattened version of listing \ref{lst:simpleleak} with immediate deallocation of temporary variables.},
+caption={Listing \ref{lst:flattened} with immediate deallocation of temporary variables.},
 captionpos=b,
@@
 \subsubsection{Decrementing reference counts}
-As discussed previously, an assignment to a variable should result in a decrement of the reference count for the object referred to by the variable before the assignment.
+An assignment to a variable should result in a decrement of the reference count for the object referred to by the variable before the assignment.
 However, this can't be done when the variable doesn't refer to anything, which is true prior to the initial assignment.
 In fact, trying to even access such a variable before it is assigned to can result in undefined results, since there is no guarantee on the contents of the variable (it could be garbage or a value from a previous stack frame).
-To prevent a call to \texttt{dec\_ref\_ctr} on a variable before the initial assignment, we initially tried to statically determine when the first assignment occurred.
+To prevent a call to \afunction{dec\_ref\_ctr} on a variable before the initial assignment, we initially tried to statically determine when the first assignment occurred.
 This approach failed when we realized that for some programs, it is not possible to determine where the initial assignment occurs at compile-time.
-Listing~\ref{lst:initialassign} shows an example program where the initial assignment to the variable \texttt{x} cannot be determined statically.
-If the input provided by the user evaluates to \texttt{True}, then the initial assignment to $x$ occurs at line 2.
-However, if the input evaluates to \texttt{False}, then the initial assignment to \texttt{x} occurs at line 3.
-If we transform the input program so that a call is made to decrement the reference count for \texttt{x} immediately prior to the assignment on line 3, then there is the possibility that this will cause the program to crash if the user provides an input that evaluates to \texttt{False}.
+Listing~\ref{lst:initialassign} shows an example program where the initial assignment to the variable \avariable{x} cannot be determined statically.
+If the input provided by the user evaluates to \aliteral{True}, then the initial assignment to $x$ occurs at line 2.
+However, if the input evaluates to \aliteral{False}, then the initial assignment to \avariable{x} occurs at line 3.
+If we transform the input program so that a call is made to decrement the reference count for \avariable{x} immediately prior to the assignment on line 3, then there is the possibility that this will cause the program to crash if the user provides an input that evaluates to \aliteral{False}.
 
 To solve this issue, we make a simple change.
-We ensure that the value of \texttt{x} is always initialized, by assigning it the value zero at the program scope where it is defined\footnote{Python defines this as the block where a variable is initially assigned to.}.
-In our compiler, this is equivalent to a primitive of type \texttt{int}, but this can be any sentinel value that results in a no-op when a call is made to \texttt{dec\_ref\_ctr}.
-By ensuring that all local variables are assigned a sentinel value at the beginning of the scope where they are defined, we can safely transform our input program to call \texttt{dec\_ref\_ctr} immediately preceding an assignment, and \texttt{inc\_ref\_ctr} immediately succeeding an assignment.
+We ensure that the value of \avariable{x} is always initialized, by assigning it the value zero at the program scope where it is defined\footnote{Python defines this as the block where a variable is initially assigned to.}.
+In our compiler, this is equivalent to a primitive of type \atype{int}, but this can be any sentinel value that results in a no-op when a call is made to \afunction{dec\_ref\_ctr}.
+By ensuring that all local variables are assigned a sentinel value at the beginning of the scope where they are defined, we can safely transform our input program to call \afunction{dec\_ref\_ctr} immediately preceding an assignment, and \afunction{inc\_ref\_ctr} immediately succeeding an assignment.
+Listing~\ref{lst:initialassign2} shows the results of this change.
 
@@
 numbers=left,
-caption={An example program where the initial assignment to \texttt{x} cannot be determined statically.},
+caption={An example program where the initial assignment to \avariable{x} cannot be determined statically.},
 captionpos=b,
@@
 
+\begin{figure}[h]
+\begin{lstlisting}[
+language=Python,
+frame=none,
+xleftmargin=0.2in,
+numbers=left,
+caption={Listing~\ref{lst:initialassign} after passing through our reference counting AST transformation. Increments/decrements to reference counts are not shown for brevity.},
+captionpos=b,
+label=lst:initialassign2
+]
+x$0 = 0
+tmp0 = 0
+tmp1 = 0
+tmp2 = 0
+tmp0 = input()
+if tmp0:
+    tmp1 = [1]
+    x$0 = tmp1
+    tmp1 = 0
+tmp0 = 0
+tmp2 = [2]
+# Here, decref(x$0) is a no-op if 
+# x$0 contains the sentinel value.
+x$0 = tmp2 
+tmp2 = 0
+\end{lstlisting}
+\end{figure}
+
 \subsubsection{Functions}
-The compiler creates temporary variables in a number of situations (namely function calls and as a result of flattened expressions).
-These temporary variables need not hang on to their memory reference past the next assignment to the actual result variable.
-Any temporary variable is dereferenced immediately at the end of the current scope if it is not the variable being returned.
-If it is the return variable we ignore decrementing the reference on this variable, thereby assuring the caller that the object returned is correctly reference counted for its assignment.
-If we had decremented the reference counter, due to the immediate nature of garbage collection, we would have been returning a value which did not point to an allocated memory region.
+There are some special provisions that need to be taken to correctly handle functions and return values from functions.
+First, the reference count for any arguments declared in the function's prototype must be incremented at the beginning of the function, and decremented at the end.
+The primary reason for this is that reassignment of an argument in the body of the function results in a decrement of the reference count.
+If we chose not to increment the reference counts for function arguments, then a reassignment in the body would result in a premature deallocation of the object, likely leading to program failures.
 
-\subsubsection{Addition of Lists}
-In the current runtime an entirely new list is created when two lists are added together.
-At the assignment level our compiler can determine (due to the flattened structure) that the left-hand-side and the right-hand-side of the addition need their reference counters manipulated, however the assignment from an addition to a list creates an entirely new list which should properly increment the reference counter to all objects in both lists.
+Return statements are also tricky to get right.
+Since return statements can appear in any order in the body of a function, we have to ensure that a return statement will not skip any decrement of reference counts associated with the end of the function scope.
+To ensure that a return statement will execute this ``cleanup'' code, we add a function context that contains the list of local assignments and function arguments.
+This context is available when processing any return nodes, and therefore the decrement of reference counts can be inserted prior to the return from the function.
 
-\subsubsection{Debugging Integration}
-To facilitate debugging in our compiler, we modify the x86 assembly generator to include calls to set up our \texttt{pymem} module and print statistics at the end of the program.
+Finally, return values must be accounted for.
+Since an object may be created in a function and returned to the caller, we cannot decrement its reference count to zero before returning from the function, otherwise it will be immediately deallocated.
+Our first heuristic to handle this removed the call to decrement the reference count for the variable being returned from a function.
+Correspondingly, the caller did not increment its reference count after assigning a return value from a function.
+However, this heuristic failed for the case where a function argument is returned.
+In this case, the function argument has one additional reference count that is introduced at the beginning of the function, and therefore will never be decremented to zero.
+Instead, we chose to implement a different approach by deferring the decrement of the reference count for any variable that is being returned from a function.
+This is done by adding the \afunction{autorelease}\footnote{autorelease is borrowed from the Objective-C keyword with the same meaning} function to the runtime library.
+This function will defer the decrement of a reference count for an object until the next call to \afunction{dec\_ref\_ctr}.
+This is done by incrementing a new field, \afield{deferred\_cnt} in the C structure representing an object.
+On the next call to \afunction{dec\_ref\_ctr}, this field is consulted to determine how many times to decrement the reference count.
+
+\subsubsection{Other challenges}
+In the runtime, a new list is created when two lists are added together.
+Our compiler correctly handles reference counting for the left hand side and the right hand side of the addition (handled by the intermediate flatten phase), however the newly created list references the individual elements from both lists.
+Therefore, the runtime had to be modified to correctly increment the reference counts for each of the elements from the original lists which are referenced in the new, concatenated list.
+
+We also faced challenges getting constructors to work with our design.
+Constructors receive a reference to the constructed object before there is a chance for the program to call \afunction{inc\_ref\_ctr} on the object.
+The result is that the constructor increments and decrements the reference count to zero, thereby prematurely deallocating the object.
+Further work is needed to fix this issue.
+
+Finally, heapification presents some unique challenges that require additional work to overcome.
+Heapification occurs after explication, and therefore after our reference counting AST transformation.
+This causes problems when the heapification phase introduces additional assignments to heapify variables.
+These additional assignments are then never referenced counted correctly.
+Also, the process of heapification encapsulates a variable in a list, and transforms assignments to that variable to assignments to a subscript.
+This is problematic because our \afunction{set\_subscript} runtime function already handles incrementing and decrementing reference counts internally, so when an assignment that is surrounded by corresponding increment and decrement calls is converted to an assignment to a subscript, these increment and decrement calls are unnecessary.
+We cannot move the reference counting AST transformation after the explication phase, as variable are not tagged with a type after this phase.
+Therefore, to solve this issue, further work needs to done investigate if heapification can be done prior to explication.
+Despite these issues, we found that heapification still works, but causes memory leaks.
 
@@
 \label{sec:results}
+
+\subsection{Correctness}
 We evaluated our reference counting implementation for correctness by using the \texttt{valgrind} utility against a simple application that uses a large number of language features.
-The program is given below.
+This program is given in listing~\ref{lst:allfeatures}.
+Without garbage collection, valgrind reports many leaks.
+With reference counting, valgrind reports only a single leak for the same program.
+The valgrind results are given in figure~\ref{fig:valgrind}.
 
+\begin{figure}[ht]
+\footnotesize
 \begin{verbatim}
-i = 0
-list = []
-while i !=  10000:
-    list = list + [[i]]
-    i = i + 1
+LEAK SUMMARY:
+   definitely lost: 560,248 bytes in 20,011 blocks
+   indirectly lost: 198,663,196 bytes in 20,001 blocks
+     possibly lost: 1,718,244 bytes in 20,035 blocks
+   still reachable: 0 bytes in 0 blocks
+        suppressed: 0 bytes in 0 blocks
 \end{verbatim}
-
-Without any garbage collection, valgrind reports
+\normalsize
 
@@
 LEAK SUMMARY:
-   definitely lost: 560,028 bytes in 20,001 blocks
-   indirectly lost: 199,004,992 bytes in 20,191 blocks
-     possibly lost: 1,375,008 bytes in 19,810 blocks
+   definitely lost: 16 bytes in 1 blocks
+   indirectly lost: 0 bytes in 0 blocks
+     possibly lost: 0 bytes in 0 blocks
+   still reachable: 352 bytes in 1 blocks
+        suppressed: 0 bytes in 0 blocks
 \end{verbatim}
 \normalsize
-
-With our reference counting for the same program
-
-\footnotesize
-\begin{verbatim}
-LEAK SUMMARY:
-   definitely lost: 0 bytes in 0 blocks
-   indirectly lost: 0 bytes in 0 blocks
-     possibly lost: 0 bytes in 0 blocks
-\end{verbatim}
-\normalsize
+\caption{Results of running \textbf{valgrind} on listing~\ref{lst:allfeatures} with and without reference counting.}
+\label{fig:valgrind}
+\end{figure}
 
@@
 
-To measure the overhead associated with reference counting, we run
+To measure the overhead associated with reference counting, we measure the average execution time of the sample program given in listing~\ref{lst:allfeatures} over ten runs.
+The results are given in figure~\ref{fig:overhead}.
+We see that our reference counting implementation imposes an overhead of nearly 650\% when compared to the compiler without memory management.
+
+Obviously, there is room for improvement in our implementation.
+One such improvement would be simple dispatch code to save a call to the runtime functions \afunction{inc\_ref\_ctr} and \afunction{dec\_ref\_ctr} when a variable contains a primitive.
+Going one step further, a type specialization stage could infer the possible types of variables using static program analysis, and could help to further eliminate this dispatch code.
+As mentioned in the related work, lazy deallocation can also help to batch deallocations and increase program throughput.
+
+\begin{figure}
+\normalsize % use normal font size in the figure
+\input{overhead.tex}
+\caption{Comparison of execution time overhead for reference counting vs. non-reference counting compiler}
+\label{fig:overhead}
+\end{figure}
 
@@
 
-\begin{figure}[h!]
+\begin{figure}
+\normalsize % use normal font size in the figure
+\input{treelatency.tex}
+\caption{Latency of \textit{dec\_ref\_ctr} for various size object graphs}
+\label{fig:treelatency}
+\end{figure}
+
+\section{Conclusions}
+\label{sec:conclusion}
+
+Reference counting is one of several options available to language designers who are looking for solutions to automated memory management.
+We provide insight into the complexity of implementing such a feature, by extending an existing Python to x86 compiler to support reference counting garbage collection.
+We find that while the premise behind reference counting is relatively simple, there are many nuances to integrating reference counting into such a compiler, especially in the presence of relatively complex language features such as closures and objects.
+
+\appendix
+% This is the text of the appendix, if you need one.
+\section{Evaluation Programs}
+
+\begin{figure}[h]
 \begin{lstlisting}[
@@
 numbers=left,
-caption={An example input program that creates a tree of objects contained in lists.},
+caption={\textit{tree.py}: creates a 4-ary tree of objects using lists.},
 captionpos=b,
@@
 
-\begin{figure}
-\normalsize % use normal font size in the figure
-\input{treelatency.tex}
-\caption{Latency of \textit{dec\_ref\_ctr} for various size object graphs}
-\label{fig:treelatency}
+\begin{figure}[h]
+\begin{lstlisting}[
+language=Python,
+frame=none,
+xleftmargin=0.2in,
+numbers=left,
+caption={\textit{allfeatures.py}: Used to evaluate correctness of our reference counting implementation by using almost all our language features.},
+captionpos=b,
+label=lst:allfeatures
+]
+a = {1:2}
+print a
+a = 0
+
+class B:
+    x = 1
+
+print B.x
+b = B()
+b.y = 2
+print b.y
+b = 0
+
+class C(B):
+    y = 2
+c = C()
+c.z = 3
+print c.x
+print c.y
+print c.z
+
+def f(x,y):
+    g = lambda x: x
+    i = 0
+    lst = []
+    while i !=  10000:
+        lst = lst + [[i]]
+        i = i + 1
+    return g(lst)
+
+print f(1,2)
+\end{lstlisting}
 \end{figure}
-
-\section{Future work}
-\label{sec:future}
-
-Optimizations.
-
-\section{Conclusions}
-\label{sec:conclusion}
-
-\subsection{Disadvantages}
-
-\subsection{Advantages}
-
-%\appendix
-%\section{Appendix Title}
-% This is the text of the appendix, if you need one.
 

### 2011-12-15 paper/paper.tex
fix some grammar

@@
 It is typical that these algorithms trade-off throughput for responsiveness.
-However, these pause times can still be an issue, especially in real time systems, where a hard guarantees on latency are needed.
+However, these pause times can still be an issue, especially in real time systems, where hard guarantees on latency are needed.
 Reference counting has been discussed as a solution to this problem since it incrementally deallocates objects at the program point where all references are destroyed, but classic implementations of reference counting can have serious performance concerns when one deallocation results in cascading deallocations\cite{boehm}.
@@
 However, this can't be done when the variable doesn't refer to anything, which is true prior to the initial assignment.
-In fact, trying to even access such a variable before it is assigned to can result in undefined results, since there is no guarantee on the contents of the variable (it could be garbage or a value from a previous stack frame).
+In fact, trying to even access such a variable before it is assigned produces undefined behavior since there is no guarantee on the contents of the variable (it could be garbage or a value from a previous stack frame).
 To prevent a call to \afunction{dec\_ref\_ctr} on a variable before the initial assignment, we initially tried to statically determine when the first assignment occurred.

### 2011-12-15 paper/paper.tex
- Edits through 3.2.4: decrementing reference counts.

@@
 Python is a modern, high level language that uses dynamic typing and is run within an interpreter.
-It manages memory for the programmer, and is considered by many to be a good ``prototyping'' language, in which programmers can quickly and easily develop an application.
+It manages memory for the programmer and is considered to be a good ``prototyping'' language, in which programmers can quickly and easily develop an application.
 Automatic memory management plays a large part in supporting this role, as it relieves the programmer from the tedious task of remembering when to deallocate memory.
@@
 There are currently two widely used strategies to perform automatic memory management.
-One such strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.
-This set of objects includes all globals, as well as the objects which are reachable from the current stack frame.
-Any objects that are traversed during this operation are considered reachable and therefore cannot be deallocated.
-Conversely, any objects not traversed are no longer reachable, and can be deallocated.
-The other primary strategy for automated memory management, is reference counting garbage collection.
+One strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.
+This set of objects includes all global objects, as well as the objects which are reachable from the current stack frame.
+Any objects that are traversed during this operation are marked as reachable and therefore cannot be deallocated.
+Any object which is not marked is a candidate for deallocation.
+The other primary strategy for automated memory management is reference counting garbage collection.
 Reference counting works by keeping a count of references to a particular object.
-References are tracked by incrementing and decrementing an object's reference count when an assignment to a pointer occurs, or when a pointer variable goes out of scope.
+References are tracked by incrementing and decrementing an object's reference count when an assignment to a pointer occurs or when a pointer variable goes out of scope.
 When the reference count reaches zero, it is safe to deallocate the object.
 
-The performance characteristics of each strategy can vary considerably, depending on implementation.
+The performance characteristics of each strategy can vary considerably depending on implementation.
 Tracing garbage collectors can sometimes be associated with long pause times, where application work must be temporarily stopped to allow the entire contents of the heap to be scanned by the garbage collector.
@@
 \label{sec:related}
-In general a reference count garbage collector will have shorter pause times for the garbage collection phase, but will incur a higher performance penalty \cite{joisha}\cite{blackburn}.
-Reference-counting collectors have been implemented which defer the deallocation, called lazy reference-counting, to some later time and some improvements on such strategies have been made \cite{boehm}.
-Without extra work a strictly reference-counting collector will leak memory when references contain cycles, and multiple algorithms have been presented as solutions to this problem.
-\cite{lin}\cite{trancon}
+In general, a reference count garbage collector will have shorter pause times for the garbage collection phase but will incur a higher performance penalty \cite{joisha}\cite{blackburn}.
+To address these performance concerns, lazy reference counting collectors have been implemented which defer deallocation to some later time.
+Improvements on such strategies are proposed by Boehm \cite{boehm}.
+In \cite{levanoni}, Levanoni, et.
+al.
+show that reference counting is a valid approach for multprocessor systems.
 
-One potential advantage that reference-counting garbage collectors have is the ability to be implemented in hardware \cite{joao}.
-Given the significant cost of garbage collection \cite{hertz} any hardware assisted acceleration would be highly desirable if it is flexible enough to adapt across multiple garbage collection techniques.
+One potential advantage that reference counting garbage collectors have is the ability to be implemented in hardware \cite{joao}.
+Given the significant cost of garbage collection \cite{hertz}, any hardware assisted acceleration would be highly desirable if it is flexible enough to adapt across multiple garbage collection techniques.
+Real-time systems are often very concerned with latency, and therefore do not typically make use of garbage collection.
+However, reference counting garbage collection has been adapted to hard real-time systems \cite{ritzaou}.
 
-Real-time systems are rarely implemented in languages which are garbage collected, but reference-counting garbage collection has been adapted to hard real-time systems \cite{ritzaou}.
-Due to immediate knowledge that an object is no longer in use reference counting garbage collection can allow for destructors or finalizers to be run immediately which can enhance the clean up of system resources.
+Because reference counting garbage collection can deallocate an object at the specific program point where there are no longer any references, it can allow for finalizers to be run immediately.
+This makes the use of finalization a valid approach to releasing system resources.
 
-In \cite{levanoni}, Levanoni, et. al. show that reference counting is a valid approach for multprocessor systems.
+One of the primary disadvantages of a strictly reference counting collector is that it will leak memory when references contain cycles.
+Multiple algorithms have been presented as solutions to this problem \cite{lin}\cite{trancon}.
 
-\par
 \section{Implementation}
@@
 
-We start with an existing compiler based upon previous work\cite{siek}.
+We start with an existing compiler based upon previous work \cite{siek}.
 This compiler converts a subset of the Python language to x86 assembly and supports one of the core features of Python, namely that everything is a first class object.
-Prior to our modifications, the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs, with minimal memory constraints.
-After our modifications, the compiler properly maintains reference counts for all objects created, and automatically deallocates objects when there are no longer any references to them.
+Prior to our modifications, the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs with minimal memory constraints.
+After our modifications, the compiler properly maintains reference counts for all objects created and automatically deallocates objects when there are no longer any references to them.
 
-Implementing reference counting requires a minimum set of steps to be taken.
+Implementing reference counting requires the following set of steps to be taken:
 
@@
 
-These steps require modifications to the architecture of our compiler, as well as the supporting runtime library responsible for creating objects and allocating memory.
-We discuss these modifications in the following sections.
+These steps require modifications to the architecture of our compiler and to the supporting runtime library responsible for creating objects and allocating memory.
+We will discuss these modifications in the following sections.
 
@@
 To support reference counts, this C structure is modified to include an int field called \afield{ref\_ctr}.
-Then all of the functions to create objects are modified to set the initial reference count to zero.
+Then, all of the functions to create objects are modified to set the initial reference count to zero.
 
@@
 In the \afunction{free\_\textit{type}} functions, it is necessary to add further calls to decrement the reference counter of any objects referenced by the object being deallocated.
-For the case of a list, we have the \afunction{free\_list} function, which iterates over each element in the list and decrements its reference count.
+In the case of a list, we have the \afunction{free\_list} function, which iterates over each element in the list and decrements its reference count.
 For a dictionary, we have the \afunction{free\_dict} function, which decrements the reference counter on keys as well as values contained in the dictionary, since both can be objects.
-Note that for any case of an object that references other objects, this can can result in deallocation of the referenced object, if the only reference to it is held by the parent object.
+Note that for any case of an object that references other objects, this can can result in deallocation of the referenced object if the only reference to it is held by the parent object.
 In the worst case, this could lead to a number of deallocations equivalent to the number of nodes in the object graph created by the parent object.
 
-Since every object in our system may contain references to another object, we modify our runtime to correctly increment the reference count when such as reference is established.
+Since every object in our system may contain references to another object, we modify our runtime to correctly increment the reference count when such a reference is established.
 For example, an assignment to a list or dictionary subscript results in a call to the \afunction{set\_subscript} runtime function.
-This function is modified to first decrement the reference count for any object that may have been previously referenced by the given subscript, and then increment the reference count for the new object that is assigned to the given subscript.
+This function is modified to first decrement the reference count for any object that may have been previously referenced by the given subscript, and to then increment the reference count for the new object that is assigned to the given subscript.
 Similarly, a closure maintains a reference to its list of free variables, and a class object maintains a reference to its base classes.
@@
 In some cases, the type definition for an object in our runtime did not aggregate a \atype{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \atype{class\_struct}).
-For this reason, significant changes were made to the runtime to ensure that any type that references another type, has a generic pointer to a \atype{big\_pyobj} structure, so that the reference counts can be correctly maintained.
-To ensure that we did not make any mistakes in this process, we create a runtime test suite that uses the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are deallocated when their reference count reaches zero.
-This runtime test suite is important to unit test the runtime library, before attempting to use it from our compiler.
+For this reason, significant changes were made to the runtime to ensure that any type that references another type has a generic pointer to a \atype{big\_pyobj} structure, so that the reference counts can be correctly maintained.
+To ensure that we did not make any mistakes in this process, we created a runtime test suite that used the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are deallocated when their reference count reaches zero.
+This runtime test suite is important to unit test the runtime library before attempting to use it from our compiler.
 
 Finally, in order to verify the correctness of our implementation, we create a set of abstractions on top of the system allocation functions \afunction{malloc} and \afunction{free}.
-These abstractions are part of our runtime and provide us the following benefits.
+These abstractions are part of our runtime and provide us the following benefits:
 
@@
 Every allocation of an object in our runtime is modified to use the \afunction{pymem\_new} and \afunction{pymem\_free} instead of \afunction{malloc} and \afunction{free}.
-As a result of this, we can determine exactly how much memory has been allocated and what objects have still not been deallocated, by calling the \afunction{pymem\_print\_statistics} function at any given program point.
+As a result, we can determine exactly how much memory has been allocated and what objects have still not been deallocated by calling the \afunction{pymem\_print\_statistics} function at any given program point.
 In practice, this is only called at the end of each program\footnote{This is achieved by modifying the x86 generation phase of our compiler to include calls to set up our \texttt{pymem} module and print statistics at the end of the program.
@@
 The caller is then returned the portion of memory immediately after this pointer.
-Subsequent calls to \afunction{pymem\_free} then have access to the linked list node via the pointer, and can update the node to mark the memory as deallocated, as well as the deallocation time.
+Subsequent calls to \afunction{pymem\_free} then have access to the statistics in the linked list node via the pointer.
+These calls will then update the node to mark the memory as deallocated and record the deallocation time.
 The memory can then be freed independently of the linked list containing the statistics.
@@
 Each stage operates on an abstract syntax tree (AST).
-The relevant new stages are the intermediate flatten stage, and the reference counting AST transformation stage.
+The relevant new stages are the intermediate flatten stage and the reference counting AST transformation stage.
 
@@
 
-Recall that in order to properly maintain reference counts, we need to increment the reference count for an object when a reference is created for it, and decrement the reference count when a reference is destroyed.
+Recall that in order to properly maintain reference counts, we need to increment the reference count for an object when a reference is created for it and decrement the reference count when a reference is destroyed.
 References are created by assigning a memory location (which contains an object) to a variable.
@@
 
-The most natural way to implement this in our compiler, is to transform our input programs AST such that any assignment to a variable is preceded by a decrement to the reference count for the object the variable refers to, and is succeeded by an increment to the reference count for the newly assigned object.
+The most natural way to implement this in our compiler is to transform our input program's AST such that any assignment to a variable is preceded by a decrement to the reference count for the object the variable refers to, and is succeeded by an increment to the reference count for the newly assigned object.
 This is the approach we took, but there are some subtleties with respect to where this operation is placed within the compiler.
 Our initial hypothesis was that we could easily add the increment and decrement reference counter operations in the existing ``Flatten AST'' stage.
-We quickly realized that because this stage is after type explication, there exist variables which are not \textit{tagged} with a type.
-Since we only want to decrement reference counts for \textit{tagged} types\footnote{Actually, we only want to modify reference counts for a subset of tagged types that only includes ``big'' types.
+We quickly realized that because this stage is after type explication, there exist variables which are not tagged with a type.
+Since we only want to decrement reference counts for tagged types\footnote{Actually, we only want to modify reference counts for a subset of tagged types that only includes ``big'' types.
 However, to simplify our implementation, we transform our AST to always call our increment and decrement runtime functions, but these functions will ignore the primitive types of \atype{int} and \atype{boolean}.}, the only option is to put this transformation stage before type explication.
-In addition to modifying the AST prior before/after an assignment, the reference counting transformation is also responsible for determining which variables are local to a given scope and decrementing the counts for these references at the end of the given program scope.
+In addition to modifying the AST before and after an assignment, the reference counting transformation is also responsible for determining which variables are local to a given scope and decrementing the counts for these references at the end of the given program scope.
 
@@
 Prior to our explicate phase, complex expressions are not flattened.
-This leads to problems where memory is leaked, since no assignments ever occur to temporary objects.
-Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between to lists to the variable \avariable{x}.
+This leads to problems where memory is leaked, since no assignments are ever made to temporary objects.
+Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between two lists to the variable \avariable{x}.
 The assignment to the variable \avariable{x} results in a corresponding increment reference count on the result of the list addition.
 However, the temporary lists, \aliteral{[1]} and \aliteral{[2]} are leaked if we don't take any additional steps.
-To overcome this problem, we introduce a intermediate flattening stage, prior to our reference counting AST transformation.
+To overcome this problem, we introduce an intermediate flattening stage prior to our reference counting AST transformation.
 This results in the corresponding code given in Listing~\ref{lst:flattened} after flattening.
@@
 This is undesirable because it changes the deallocation behavior of the program from the expected behavior.
-For original input program in listing~\ref{lst:simpleleak}, the expectation is that when the variable \avariable{x} is assigned the value zero, the reference count for the list object should be decremented to zero and the object deallocated.
+For the original input program in listing~\ref{lst:simpleleak}, the expectation is that when the variable \avariable{x} is assigned the value zero, the reference count for the list object should be decremented to zero and the object deallocated.
 However, in the flattened program, this is no longer the case since the \avariable{tmp2} variable continues to reference the object.

### 2011-12-15 paper/paper.tex
tweak grammar and commas

@@
 One strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.
-This set of objects includes all global objects, as well as the objects which are reachable from the current stack frame.
+This set of objects includes all global objects as well as the objects which are reachable from the current stack frame.
 Any objects that are traversed during this operation are marked as reachable and therefore cannot be deallocated.
@@
 
-Because reference counting garbage collection can deallocate an object at the specific program point where there are no longer any references, it can allow for finalizers to be run immediately.
+Because reference counting garbage collection can deallocate an object at the specific program point where there are no longer any references it can allow for finalizers to be run immediately.
 This makes the use of finalization a valid approach to releasing system resources.
@@
 This compiler converts a subset of the Python language to x86 assembly and supports one of the core features of Python, namely that everything is a first class object.
-Prior to our modifications, the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs with minimal memory constraints.
+Prior to our modifications the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs with minimal memory constraints.
 After our modifications, the compiler properly maintains reference counts for all objects created and automatically deallocates objects when there are no longer any references to them.
@@
 \item Reference counts must be decremented when a program reaches the end of a given scope, since local variables are destroyed.
-\item When an object's reference count reaches zero, it is no longer reachable and can be safely destroyed.
+\item When an object's reference count reaches zero it is no longer reachable and can be safely destroyed.
 
@@
 According to the steps for reference counting mentioned previously, we need to deallocate an object whenever its reference count reaches zero.
-To provide this behavior, the \afunction{dec\_ref\_ctr} function has to check if the reference count has reached zero, and if so, delegate to a corresponding \afunction{free\_\textit{type}} method to free an object of the given type.
+To provide this behavior, the \afunction{dec\_ref\_ctr} function has to check if the reference count has reached zero and, if so, delegate to a corresponding \afunction{free\_\textit{type}} method to free an object of the given type.
 
@@
 
-In some cases, the type definition for an object in our runtime did not aggregate a \atype{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \atype{class\_struct}).
-For this reason, significant changes were made to the runtime to ensure that any type that references another type has a generic pointer to a \atype{big\_pyobj} structure, so that the reference counts can be correctly maintained.
-To ensure that we did not make any mistakes in this process, we created a runtime test suite that used the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are deallocated when their reference count reaches zero.
+In some cases the type definition for an object in our runtime did not aggregate a \atype{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \atype{class\_struct}).
+For this reason significant changes were made to the runtime to ensure that any type that references another type has a generic pointer to a \atype{big\_pyobj} structure, so that the reference counts can be correctly maintained.
+To ensure that we did not make any mistakes in this process we created a runtime test suite that used the abstractions provided by the runtime library to create objects, increment and decrement their reference counts, and ensure that all objects are deallocated when their reference count reaches zero.
 This runtime test suite is important to unit test the runtime library before attempting to use it from our compiler.
 
-Finally, in order to verify the correctness of our implementation, we create a set of abstractions on top of the system allocation functions \afunction{malloc} and \afunction{free}.
+Finally, in order to verify the correctness of our implementation we create a set of abstractions on top of the system allocation functions \afunction{malloc} and \afunction{free}.
 These abstractions are part of our runtime and provide us the following benefits:
@@
 As a result, we can determine exactly how much memory has been allocated and what objects have still not been deallocated by calling the \afunction{pymem\_print\_statistics} function at any given program point.
-In practice, this is only called at the end of each program\footnote{This is achieved by modifying the x86 generation phase of our compiler to include calls to set up our \texttt{pymem} module and print statistics at the end of the program.
+In practice this is only called at the end of each program\footnote{This is achieved by modifying the x86 generation phase of our compiler to include calls to set up our \texttt{pymem} module and print statistics at the end of the program.
 } to ensure that all memory has been deallocated.
@@
 An assignment to a variable should result in a decrement of the reference count for the object referred to by the variable before the assignment.
-However, this can't be done when the variable doesn't refer to anything, which is true prior to the initial assignment.
+This cannot, however, be done when the variable does not refer to an allocated memory region, which is true prior to the initial assignment.
 In fact, trying to even access such a variable before it is assigned produces undefined behavior since there is no guarantee on the contents of the variable (it could be garbage or a value from a previous stack frame).
 To prevent a call to \afunction{dec\_ref\_ctr} on a variable before the initial assignment, we initially tried to statically determine when the first assignment occurred.
-This approach failed when we realized that for some programs, it is not possible to determine where the initial assignment occurs at compile-time.
+This approach failed when we realized that for some programs it is not possible to determine where the initial assignment occurs at compile-time.
 Listing~\ref{lst:initialassign} shows an example program where the initial assignment to the variable \avariable{x} cannot be determined statically.
@@
 
-To solve this issue, we make a simple change.
-We ensure that the value of \avariable{x} is always initialized, by assigning it the value zero at the program scope where it is defined\footnote{Python defines this as the block where a variable is initially assigned to.}.
+To solve this issue it was necessary to make a simple change: ensure that the value of \avariable{x} is always initialized by assigning it the value zero at the program scope where it is defined\footnote{Python defines this as the block where a variable is initially assigned to.}.
 In our compiler, this is equivalent to a primitive of type \atype{int}, but this can be any sentinel value that results in a no-op when a call is made to \afunction{dec\_ref\_ctr}.
@@
 
-This solution is somewhat controversial, since it can alter program behavior when the program tries to read from a variable that has not yet been assigned to.
+This solution is not ideal since it can alter program behavior when the program tries to read from a variable that has not yet been assigned a value.
 The standard Python interpreter throws a \textit{NameError} in this case, but our compiler will allow such a program to be compiled.
@@
 
-Return statements are also tricky to get right.
-Since return statements can appear in any order in the body of a function, we have to ensure that a return statement will not skip any decrement of reference counts associated with the end of the function scope.
-To ensure that a return statement will execute this ``cleanup'' code, we add a function context that contains the list of local assignments and function arguments.
+Second, \afunction{return} statements are also tricky to get right.
+Since return statements can appear in any order in the body of a function, we have to ensure that a \afunction{return} statement will not skip any decrement of reference counts associated with the end of the function scope.
+To ensure that a \afunction{return} statement will execute this ``cleanup'' code, we add a function context that contains the list of local assignments and function arguments.
 This context is available when processing any return nodes, and therefore the decrement of reference counts can be inserted prior to the return from the function.
@@
 Finally, return values must be accounted for.
-Since an object may be created in a function and returned to the caller, we cannot decrement its reference count to zero before returning from the function, otherwise it will be immediately deallocated.
+Since an object may be created in a function and returned to the caller we cannot decrement its reference count to zero before returning from the function, otherwise it will be immediately deallocated.
 Our first heuristic to handle this removed the call to decrement the reference count for the variable being returned from a function.
@@
 This is done by incrementing a new field, \afield{deferred\_cnt} in the C structure representing an object.
-On the next call to \afunction{dec\_ref\_ctr}, this field is consulted to determine how many times to decrement the reference count.
+On the next call to \afunction{dec\_ref\_ctr} this field is consulted to determine how many times to decrement the reference count.
 
 \subsubsection{Other challenges}
-In the runtime, a new list is created when two lists are added together.
+In the runtime a new list is created when two lists are added together.
 Our compiler correctly handles reference counting for the left hand side and the right hand side of the addition (handled by the intermediate flatten phase), however the newly created list references the individual elements from both lists.
@@
 Also, the process of heapification encapsulates a variable in a list, and transforms assignments to that variable to assignments to a subscript.
-This is problematic because our \afunction{set\_subscript} runtime function already handles incrementing and decrementing reference counts internally, so when an assignment that is surrounded by corresponding increment and decrement calls is converted to an assignment to a subscript, these increment and decrement calls are unnecessary.
+This is problematic because our \afunction{set\_subscript} runtime function already handles incrementing and decrementing reference counts internally; so when an assignment that is surrounded by corresponding increment and decrement calls is converted to an assignment to a subscript these increment and decrement calls are redundant.
 We cannot move the reference counting AST transformation after the explication phase, as variable are not tagged with a type after this phase.
 Therefore, to solve this issue, further work needs to done investigate if heapification can be done prior to explication.
-Despite these issues, we found that heapification still works, but causes memory leaks.
+Despite these issues we found that heapification still works, but may cause memory leaks.
 
@@
 This program is given in listing~\ref{lst:allfeatures}.
-Without garbage collection, valgrind reports many leaks.
-With reference counting, valgrind reports only a single leak for the same program.
-The valgrind results are given in figure~\ref{fig:valgrind}.
+Without garbage collection \texttt{valgrind} reports many memory leaks.
+With reference counting \texttt{valgrind} reports only a single leak for the same program.
+The \texttt{valgrind} results are given in figure~\ref{fig:valgrind}.
 
@@
 
-To measure the overhead associated with reference counting, we measure the average execution time of the sample program given in listing~\ref{lst:allfeatures} over ten runs.
+To measure the overhead associated with reference counting we measure the average execution time of the sample program given in listing~\ref{lst:allfeatures} over ten runs.
 The results are given in figure~\ref{fig:overhead}.
@@
 
-Obviously, there is room for improvement in our implementation.
+There is obvious room for improvement in our implementation.
 One such improvement would be simple dispatch code to save a call to the runtime functions \afunction{inc\_ref\_ctr} and \afunction{dec\_ref\_ctr} when a variable contains a primitive.
@@
 The latency introduced by the \textit{dec\_ref\_ctr} is linear with the number of nodes in the object graph being deallocated.
-To illustrate this, the program given in listing~\ref{lst:tree} is run with different inputs, where the input is the height of a 4-ary tree.
+To illustrate this the program given in listing~\ref{lst:tree} is run with different inputs, where the input is the height of a 4-ary tree.
 The number of nodes in such a tree is given by $\frac{{k}^{\left(h+1\right)}-1}{k-1}$ where $k=4$, and $h$ is the height of tree.
@@
 Reference counting is one of several options available to language designers who are looking for solutions to automated memory management.
-We provide insight into the complexity of implementing such a feature, by extending an existing Python to x86 compiler to support reference counting garbage collection.
+We provide insight into the complexity of implementing such a feature by extending an existing Python to x86 compiler to support reference counting garbage collection.
 We find that while the premise behind reference counting is relatively simple, there are many nuances to integrating reference counting into such a compiler, especially in the presence of relatively complex language features such as closures and objects.
